---
title: "Getting Started with psvr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with psvr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

Classical SVR minimises absolute-error losses (MAE, MSE), which are
misaligned with the scale-free accuracy criteria standard in forecasting.
An error of 1 unit is negligible when the target is 1 000 but large when
it is 2.

**psvr** implements four SVR variants derived from percentage-error loss
functions (Benavides-Herrera et al., 2026), accessed through one fitter per
model family:

| Model | Function | `sym_type` | Solver           |
|-------|----------|------------|------------------|
| 3     | `psvr_rmspe()` | `"none"` | linear system    |
| 4     | `psvr_rmspe()` | `"even"` / `"odd"` | linear system |
| 1     | `psvr_mape()`  | `"none"` | quadratic program |
| 2     | `psvr_mape()`  | `"even"` / `"odd"` | quadratic program |

`sym_type = "even"` enforces even symmetry `f(x) = f(-x)`; `"odd"` enforces
odd symmetry. Use the symmetric variants only with kernels that satisfy
Assumption 3 of the paper (RBF and even-degree polynomial kernels do).

These are the two direct fitters. They replaced the unified `psvr()` entry
point in 0.0.2.9012, which had in turn replaced four separate wrappers in
0.0.2.9010: seven of `psvr()`'s eleven arguments were conditional on which
family you were fitting, so the families are now separate functions. Both
return the same classes the tidymodels engine returns, so a fit obtained
either way behaves identically.

All models require **strictly positive targets** (`y > 0`), which is the
condition under which percentage residuals are well-defined.

## Boston Housing data

We reproduce the experiments reported in Tables 1–2 of the paper using
the Boston Housing dataset (Harrison & Rubinfeld, 1978). All 506
observations have strictly positive median home values (MEDV, in
thousands of USD), making MAPE well-defined throughout.

```{r data}
library(psvr)
library(ggplot2)

# Boston Housing is available in the MASS package
data("Boston", package = "MASS")

# Target: median home value (all > 0)
y_all <- Boston$medv
X_raw <- as.matrix(Boston[, setdiff(names(Boston), "medv")])

stopifnot(all(y_all > 0))
cat("N =", nrow(X_raw), "  p =", ncol(X_raw),
    "  y range: [", round(min(y_all), 1), ",", round(max(y_all), 1), "]\n")
```

### 70 / 30 train–test split

Features are standardised using training-set statistics so that the RBF
kernel operates on a comparable scale across all 13 predictors.

```{r split}
set.seed(42)
n      <- nrow(X_raw)
tr_idx <- sample(n, floor(0.7 * n))

X_raw_tr <- X_raw[tr_idx, ];  y_tr <- y_all[tr_idx]
X_raw_te <- X_raw[-tr_idx, ]; y_te <- y_all[-tr_idx]

# Standardise: centre and scale by training mean/sd
col_mean <- colMeans(X_raw_tr)
col_sd   <- apply(X_raw_tr, 2, sd)
X_tr <- scale(X_raw_tr, center = col_mean, scale = col_sd)
X_te <- scale(X_raw_te, center = col_mean, scale = col_sd)
```

### Helper metrics

```{r metrics}
mape  <- function(y, yhat) mean(abs(y - yhat) / y) * 100
rmspe <- function(y, yhat) sqrt(mean(((y - yhat) / y)^2)) * 100
r2    <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)
```

### Baseline: linear regression

```{r baseline}
lm_df_tr <- as.data.frame(X_tr)
lm_df_te <- as.data.frame(X_te)
lm_fit   <- lm(y_tr ~ ., data = lm_df_tr)
lm_pred  <- predict(lm_fit, newdata = lm_df_te)

cat(sprintf("Linear regression — MAPE: %.2f%%  RMSPE: %.2f%%  R²: %.4f\n",
            mape(y_te, lm_pred), rmspe(y_te, lm_pred), r2(y_te, lm_pred)))
```

## Model 3: LS-SVR with RMSPE

The LS-SVR formulation replaces the QP with a linear system by using a
quadratic penalty on percentage residuals. The dual reduces to:

$$
\begin{bmatrix} 0 & \mathbf{1}^\top \\ \mathbf{1} & \Omega + Y_\Gamma \end{bmatrix}
\begin{bmatrix} b \\ \boldsymbol{\alpha} \end{bmatrix}
= \begin{bmatrix} 0 \\ \mathbf{y} \end{bmatrix}
$$

where $Y_\Gamma = \operatorname{diag}(y_1^2/\Gamma, \ldots, y_N^2/\Gamma)$.

```{r rmspe-lssvr}
# make_kernel() returns a closure K(xi, xj) = exp(-||xi - xj||^2 / (2 sigma^2))
K <- make_kernel("rbf", sigma = 1)

# gamma = 5000: regularisation; larger gamma -> smaller Y_Gamma diagonal -> tighter fit
fit_ls <- psvr_rmspe(X_tr, y_tr, kernel = K, gamma = 5000)
pred_ls <- predict(fit_ls, X_te)

cat(sprintf("LS-SVR RMSPE  — MAPE: %.2f%%  RMSPE: %.2f%%  R²: %.4f\n",
            mape(y_te, pred_ls), rmspe(y_te, pred_ls), r2(y_te, pred_ls)))
print(fit_ls)
```

```{r rmspe-plot, echo = FALSE}
lim <- range(c(y_te, pred_ls))
data.frame(actual = y_te, predicted = pred_ls) |>
  ggplot(aes(actual, predicted)) +
  geom_point(colour = "#2c7bb6", alpha = 0.65, size = 1.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  coord_equal(xlim = lim, ylim = lim) +
  labs(x = "Actual MEDV", y = "Predicted MEDV",
       title = "Model 3: LS-SVR with RMSPE")
```

```{r rmspe-coef}
cf_ls <- coef(fit_ls)
# alpha:        N dual variables; weight each training point's kernel
#               contribution in f(x) = sum_k alpha_k K(x_k, x) + b
#               (all N points, no sparsity)
# b:            bias / intercept term
# support_data: all N training inputs stored for prediction
cat(sprintf("b = %.4f  |  alpha range: [%.4f, %.4f]\n",
            cf_ls$b, min(cf_ls$alpha), max(cf_ls$alpha)))
```

## Model 1: ε-SVR with MAPE

The ε-SVR formulation optimises a QP with **sample-dependent box
constraints** $|\beta_k| \le 100C/y_k$: tighter bounds for small
targets, concentrating model capacity on low-magnitude observations.

```{r mape-svr}
# C = 10: per-sample box bound |beta_k| <= 100*C/y_k; eps = 1: tube width (% of y_k)
fit_ep <- psvr_mape(X_tr, y_tr, kernel = K, C = 10, eps = 1)
pred_ep <- predict(fit_ep, X_te)

cat(sprintf("ε-SVR MAPE    — MAPE: %.2f%%  RMSPE: %.2f%%  R²: %.4f\n",
            mape(y_te, pred_ep), rmspe(y_te, pred_ep), r2(y_te, pred_ep)))
cat(sprintf("Support vectors: %d / %d\n", length(fit_ep$beta), fit_ep$n_train))
print(fit_ep)
```

```{r mape-plot, echo = FALSE}
lim <- range(c(y_te, pred_ep))
data.frame(actual = y_te, predicted = pred_ep) |>
  ggplot(aes(actual, predicted)) +
  geom_point(colour = "#d7191c", alpha = 0.65, size = 1.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  coord_equal(xlim = lim, ylim = lim) +
  labs(x = "Actual MEDV", y = "Predicted MEDV",
       title = "Model 1: \u03b5-SVR with MAPE")
```

```{r mape-coef}
cf_ep <- coef(fit_ep)
# alpha, alpha_star: length-N dual variables (paired); the pre-pruning
#               solution.  Useful as a warm-start across CV folds; for
#               prediction use `beta` instead.
# beta:         beta_k = alpha_k - alpha_k* for each SUPPORT VECTOR only
#               (non-zero only for training points outside the
#               percentage-error ε-tube — sparse)
# b:            bias / intercept term
# support_data: training rows corresponding to support vectors only
cat(sprintf("b = %.4f  |  beta range: [%.4f, %.4f]\n",
            cf_ep$b, min(cf_ep$beta), max(cf_ep$beta)))
```

## Comparing objectives

```{r comparison}
results <- data.frame(
  Model = c("Linear regression", "LS-SVR RMSPE (Model 3)",
            "\u03b5-SVR MAPE (Model 1)"),
  MAPE  = c(mape(y_te,  lm_pred),
            mape(y_te,  pred_ls),
            mape(y_te,  pred_ep)),
  RMSPE = c(rmspe(y_te, lm_pred),
            rmspe(y_te, pred_ls),
            rmspe(y_te, pred_ep)),
  R2    = c(r2(y_te,    lm_pred),
            r2(y_te,    pred_ls),
            r2(y_te,    pred_ep))
)
results[, 2:4] <- round(results[, 2:4], 2)
knitr::kable(results, col.names = c("Model", "MAPE (%)", "RMSPE (%)", "R²"),
             align = "lrrr",
             caption = paste("Test-set performance on Boston Housing",
                             "(70/30 split, RBF kernel, single run)."))
```

Both psvr models improve on the linear baseline under their respective
percentage-error objectives. The LS-SVR formulation (Model 3) minimises
RMSPE directly and achieves the lowest squared-percentage error; the
ε-SVR formulation (Model 1) targets MAPE through sample-dependent box
constraints on the dual variables. These results are consistent with
Tables 1–2 of Benavides-Herrera et al. (2026), where SVR-MAPE reduces
test-set MAPE from 15.23% (classical ε-SVR) to around 10% relative to a
linear baseline of 17.85%.

## Using psvr with tidymodels

All four models are registered as parsnip engines and integrate
seamlessly with the tidymodels ecosystem. This enables hyperparameter
tuning via `tune_grid()`, resampling via `rsample`, and unified model
comparison via `workflow_set()`.

See the [tidymodels workflow](tidymodels-workflow.html) article for a
complete example using `psvr_rmspe_rbf()` with `tune_grid()` and
data-driven hyperparameter ranges via `rbf_sigma_psvr_data()`. The
[When to Use Percentage-Error SVR](https://pbenavidesh.github.io/psvr/articles/when-to-use-psvr.html)
article, on the package website rather than in the installed package,
shows a full `workflow_set()` comparison of all four models against
standard baselines.

## Hyperparameter search ranges

The parsnip specs register **static** default ranges — a `dials` parameter
object has to exist before any data does. Three of them are deliberately
conservative, one of them is actively wrong for LS-SVR, and one searches a
level you probably do not want, so tuning a psvr model means supplying your own
ranges through `param_info`.

Nothing below fits a model. It builds the parameter set and prints it, before
and after.

```{r ranges-defaults, message = FALSE}
library(parsnip)
library(tune)

show_ranges <- function(ps) {
  for (i in seq_len(nrow(ps))) {
    ob <- ps$object[[i]]
    cat(sprintf("  %-10s %s\n", ps$id[i],
      if (inherits(ob, "quant_param"))
        sprintf("[%s] on the %s scale",
                paste(signif(unlist(ob$range), 4), collapse = ", "),
                if (is.null(ob$trans)) "identity" else ob$trans$name)
      else sprintf("{%s}", paste(ob$values, collapse = ", "))))
  }
  invisible(ps)
}

spec_mape <- psvr_mape_rbf(cost = tune(), margin = tune(),
                           rbf_sigma = tune(), sym_type = tune()) |>
  set_engine("psvr")

extract_parameter_set_dials(spec_mape) |> show_ranges()
```

For an ε-SVR (Models 1–2) only `rbf_sigma` really needs replacing; the `cost`
and `margin` defaults are usable as they stand.

```{r ranges-mape}
extract_parameter_set_dials(spec_mape) |>
  update(
    cost      = cost_psvr(),                      # [-2, 10] log2 — fine for C
    margin    = margin_percentage(),              # 1-20% of each target
    rbf_sigma = rbf_sigma_psvr_data(X_tr),        # data-driven; see below
    sym_type  = sym_type_param(c("even", "odd"))  # drops "none"; see below
  ) |>
  show_ranges()
```

For an LS-SVR (Models 3–4) `cost` is $\Gamma$, and there the registered default
is **not** usable.

```{r ranges-rmspe}
spec_ls <- psvr_rmspe_rbf(cost = tune(), rbf_sigma = tune(),
                          sym_type = tune()) |>
  set_engine("psvr")

extract_parameter_set_dials(spec_ls) |>
  update(
    cost      = cost_psvr_ls_data(y_tr),
    rbf_sigma = rbf_sigma_psvr_data(X_tr),
    sym_type  = sym_type_param(c("even", "odd"))
  ) |>
  show_ranges()
```

No single spec carries all four helpers: `margin` exists only on the MAPE specs,
and `cost_psvr_ls_data()` applies only to the RMSPE ones.

### `rbf_sigma` is a length scale, not a precision

Of everything on this page, this is the one most likely to cost you a day.
psvr's RBF kernel is

$$K(\mathbf{x}_i, \mathbf{x}_j) =
  \exp\left(-\frac{\|\mathbf{x}_i - \mathbf{x}_j\|^2}{2\sigma^2}\right)$$

so `rbf_sigma` is a **length**, measured in the units of the preprocessed
feature space, and *larger* means a *wider* kernel. `kernlab::rbfdot(sigma =)`,
`e1071::svm(gamma =)` and `parsnip::svm_rbf(rbf_sigma =)` all use the
reciprocal convention, $\exp(-\sigma\|\cdot\|^2)$, where larger means
*narrower*. Carrying a tuned value across from one of those gives a silently
wrong kernel width — no error, no warning, just a worse model.

### `rbf_sigma_psvr_data()` needs preprocessed predictors

`rbf_sigma_psvr_data(X)` centres its range on the median pairwise Euclidean
distance between the rows of `X`, spanning one decade either side of it on the
log10 scale. Being a distance-based heuristic, it means nothing except on the
scale the model is actually fitted on: pass the standardised predictors (`X_tr`
here, or the baked output of a recipe), never the raw ones.

```{r ranges-sigma}
sigma_heuristic(X_tr)   # the geometric centre of the range printed above
```

Above `sample_size` rows (default 500) the median is taken on a random
subsample, so the centre becomes an estimate rather than the exact median — and
a seed-dependent one. Pass `seed` if you need it reproducible.

### LS-SVR needs `cost_psvr_ls_data()`, and it cannot be automated

`cost` maps to `C` on the ε-SVR models and to $\Gamma$ on the LS-SVR ones, but
both register the same default of $[-2, 10]$ on the log2 scale, i.e.
$\Gamma \le 1024$. That is the ε-SVR range. The LS-SVR optimum scales with
`var(y) * n`, which on Boston puts it near $10^5$, so a grid over the default is
**boundary-trapped** — every candidate is legal, the search reports a plausible
number, and the optimum was never inside the range. The
[tidymodels workflow](tidymodels-workflow.html) article shows the size of the
effect on a smaller dataset.

`cost_psvr_ls_data(y)` sets the ceiling from `var(y) * n`. It has to be passed
by hand: `tune` finalises parameters from the molded **predictors** alone and
never passes the outcome to `dials::finalize()`, so no machinery could compute a
`var(y)`-based bound on your behalf.

### Tuning `sym_type` cannot answer "does symmetry help"

`sym_type` registers all three levels, `{none, even, odd}`, and `"none"` *is*
the asymmetric model. So `sym_type = tune()` puts Models 1 and 2 — or 3 and 4 —
inside a single search space, and `select_best()` returns whichever candidate
won. That is a selection, not a comparison. To contrast the two families, fix
`sym_type` on two specs and compare their resampled metrics. Restricting to
`c("even", "odd")`, as above, tunes *within* the symmetric family, which is a
third question again.

### The outcome stays in original units

`recipes::step_normalize()` applies to predictors. Leave the outcome alone: the
percentage-error losses divide by `y`, so they need `y > 0` on the original
scale, and centring the outcome would destroy positivity and change what
"percentage error" even refers to. This is a genuine departure from classical
SVR, where rescaling the outcome is routine and harmless.

## References

Harrison, D. & Rubinfeld, D. L. (1978). Hedonic prices and the demand
for clean air. *Journal of Environmental Economics and Management*,
**5**(1), 81–102.

Benavides-Herrera, P., Álvarez, G., Ruiz-Cruz, R., &
Sánchez-Torres, J. D. (2026). A unified family of percentage-error
support vector regression models with symmetric kernel extensions.
*Mathematics*, 14(10), 1679.
<https://doi.org/10.3390/math14101679>
