---
title: "Getting Started with psvr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with psvr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6,
  fig.height = 4
)
```

## Introduction

Classical SVR minimises absolute-error losses (MAE, MSE), which are
misaligned with the scale-free accuracy criteria standard in forecasting.
An error of 1 unit is negligible when the target is 1 000 but large when
it is 2.

**psvr** implements four SVR variants derived from percentage-error loss
functions (Benavides-Herrera et al., 2026):

| Model | Loss | Solver |
|-------|------|--------|
| `rmspe_lssvr()` | RMSPE — least-squares | linear system |
| `rmspe_sym_lssvr()` | RMSPE — symmetric kernel | linear system |
| `mape_svr()` | MAPE — ε-insensitive | quadratic program |
| `mape_sym_svr()` | MAPE — symmetric kernel | quadratic program |

All models require **strictly positive targets** (`y > 0`), which is the
condition under which percentage residuals are well-defined.

## Installation

```{r install, eval = FALSE}
# CRAN (once released)
install.packages("psvr")

# Development version from GitHub
remotes::install_github("pbenavidesh/psvr")
```

## Boston Housing data

We reproduce the experiments reported in Tables 1–2 of the paper using
the Boston Housing dataset (Harrison & Rubinfeld, 1978). All 506
observations have strictly positive median home values (MEDV, in
thousands of USD), making MAPE well-defined throughout.

```{r data}
library(psvr)

# Boston Housing is available in the MASS package
data("Boston", package = "MASS")

# Target: median home value (all > 0)
y_all <- Boston$medv
X_raw <- as.matrix(Boston[, setdiff(names(Boston), "medv")])

stopifnot(all(y_all > 0))
cat("N =", nrow(X_raw), "  p =", ncol(X_raw),
    "  y range: [", round(min(y_all), 1), ",", round(max(y_all), 1), "]\n")
```

### 70 / 30 train–test split

Features are standardised using training-set statistics so that the RBF
kernel operates on a comparable scale across all 13 predictors.

```{r split}
set.seed(42)
n      <- nrow(X_raw)
tr_idx <- sample(n, floor(0.7 * n))

X_raw_tr <- X_raw[tr_idx, ];  y_tr <- y_all[tr_idx]
X_raw_te <- X_raw[-tr_idx, ]; y_te <- y_all[-tr_idx]

# Standardise: centre and scale by training mean/sd
col_mean <- colMeans(X_raw_tr)
col_sd   <- apply(X_raw_tr, 2, sd)
X_tr <- scale(X_raw_tr, center = col_mean, scale = col_sd)
X_te <- scale(X_raw_te, center = col_mean, scale = col_sd)
```

### Helper metrics

```{r metrics}
mape  <- function(y, yhat) mean(abs(y - yhat) / y) * 100
rmspe <- function(y, yhat) sqrt(mean(((y - yhat) / y)^2)) * 100
r2    <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)
```

### Baseline: linear regression

```{r baseline}
lm_df_tr <- as.data.frame(X_tr)
lm_df_te <- as.data.frame(X_te)
lm_fit   <- lm(y_tr ~ ., data = lm_df_tr)
lm_pred  <- predict(lm_fit, newdata = lm_df_te)

cat(sprintf("Linear regression — MAPE: %.2f%%  RMSPE: %.2f%%  R²: %.4f\n",
            mape(y_te, lm_pred), rmspe(y_te, lm_pred), r2(y_te, lm_pred)))
```

## Model 3: LS-SVR with RMSPE

The LS-SVR formulation replaces the QP with a linear system by using a
quadratic penalty on percentage residuals. The dual reduces to:

$$
\begin{bmatrix} 0 & \mathbf{1}^\top \\ \mathbf{1} & \Omega + Y_\Gamma \end{bmatrix}
\begin{bmatrix} b \\ \boldsymbol{\alpha} \end{bmatrix}
= \begin{bmatrix} 0 \\ \mathbf{y} \end{bmatrix}
$$

where $Y_\Gamma = \operatorname{diag}(y_1^2/\Gamma, \ldots, y_N^2/\Gamma)$.

```{r rmspe-lssvr}
K <- make_kernel("rbf", sigma = 1)

fit_ls <- rmspe_lssvr(X_tr, y_tr, kernel = K, gamma = 5000)
pred_ls <- predict(fit_ls, X_te)

cat(sprintf("LS-SVR RMSPE  — MAPE: %.2f%%  RMSPE: %.2f%%  R²: %.4f\n",
            mape(y_te, pred_ls), rmspe(y_te, pred_ls), r2(y_te, pred_ls)))
```

```{r rmspe-plot, echo = FALSE}
lim <- range(c(y_te, pred_ls))
plot(y_te, pred_ls,
     xlab = "Actual MEDV", ylab = "Predicted MEDV",
     main = "Model 3: LS-SVR with RMSPE",
     pch = 16, col = "#2c7bb6aa",
     xlim = lim, ylim = lim)
abline(0, 1, lty = 2, col = "grey40")
```

## Model 1: ε-SVR with MAPE

The ε-SVR formulation optimises a QP with **sample-dependent box
constraints** $|\beta_k| \le 100C/y_k$: tighter bounds for small
targets, concentrating model capacity on low-magnitude observations.

```{r mape-svr}
fit_ep <- mape_svr(X_tr, y_tr, kernel = K, C = 10, eps = 1)
pred_ep <- predict(fit_ep, X_te)

cat(sprintf("ε-SVR MAPE    — MAPE: %.2f%%  RMSPE: %.2f%%  R²: %.4f\n",
            mape(y_te, pred_ep), rmspe(y_te, pred_ep), r2(y_te, pred_ep)))
cat(sprintf("Support vectors: %d / %d\n", length(fit_ep$beta), nrow(X_tr)))
```

```{r mape-plot, echo = FALSE}
lim <- range(c(y_te, pred_ep))
plot(y_te, pred_ep,
     xlab = "Actual MEDV", ylab = "Predicted MEDV",
     main = "Model 1: \u03b5-SVR with MAPE",
     pch = 16, col = "#d7191caa",
     xlim = lim, ylim = lim)
abline(0, 1, lty = 2, col = "grey40")
```

## Comparing objectives

```{r comparison}
results <- data.frame(
  Model = c("Linear regression", "LS-SVR RMSPE (Model 3)",
            "\u03b5-SVR MAPE (Model 1)"),
  MAPE  = c(mape(y_te,  lm_pred),
            mape(y_te,  pred_ls),
            mape(y_te,  pred_ep)),
  RMSPE = c(rmspe(y_te, lm_pred),
            rmspe(y_te, pred_ls),
            rmspe(y_te, pred_ep)),
  R2    = c(r2(y_te,    lm_pred),
            r2(y_te,    pred_ls),
            r2(y_te,    pred_ep))
)
results[, 2:4] <- round(results[, 2:4], 2)
knitr::kable(results, col.names = c("Model", "MAPE (%)", "RMSPE (%)", "R²"),
             align = "lrrr",
             caption = paste("Test-set performance on Boston Housing",
                             "(70/30 split, RBF kernel, single run)."))
```

Both psvr models improve on the linear baseline under their respective
percentage-error objectives. The LS-SVR formulation (Model 3) minimises
RMSPE directly and achieves the lowest squared-percentage error; the
ε-SVR formulation (Model 1) targets MAPE through sample-dependent box
constraints on the dual variables. These results are consistent with
Tables 1–2 of Benavides-Herrera et al. (2026), where SVR-MAPE reduces
test-set MAPE from 15.23% (classical ε-SVR) to around 10% relative to a
linear baseline of 17.85%.

## References

Harrison, D. & Rubinfeld, D. L. (1978). Hedonic prices and the demand
for clean air. *Journal of Environmental Economics and Management*,
**5**(1), 81–102.

Benavides-Herrera, P., Álvarez-Álvarez, G., Ruiz-Cruz, R., &
Sánchez-Torres, J. D. (2026). A unified family of percentage-error
support vector regression models with symmetric kernel extensions.
*Mathematics*, MDPI. (under review)
