---
title: "tidymodels workflow with psvr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidymodels workflow with psvr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
```

This vignette demonstrates the full tidymodels pipeline with **psvr**: data
splitting, preprocessing, hyperparameter tuning by cross-validation, and final
model evaluation.  We use `psvr_rmspe_rbf()` (LS-SVR with RMSPE loss, RBF
kernel) and tune the regularisation parameter `cost` ($\Gamma$) against MAPE.

```{r libs}
library(psvr)
library(parsnip)
library(rsample)
library(recipes)
library(workflows)
library(tune)
library(dials)
library(yardstick)
```

## Data

The synthetic even-function dataset from the package README:
$y = 2 + x_1^2 + 0.5\,x_2^2 + \varepsilon$, $\varepsilon \sim \mathcal{N}(0,\,0.1^2)$.
Targets are strictly positive by construction ($y > 0$).

```{r data}
set.seed(42)
n   <- 200
x1  <- runif(n, -3, 3)
x2  <- runif(n, -3, 3)
y   <- 2 + x1^2 + 0.5 * x2^2 + rnorm(n, sd = 0.1)
dat <- data.frame(y = y, x1 = x1, x2 = x2)
```

## 1 — Split

```{r split}
set.seed(1)
split <- initial_split(dat, prop = 0.75)
train <- training(split)
test  <- testing(split)
```

## 2 — Preprocessing recipe

Centre and scale all predictors so the RBF kernel operates on a
standardised feature space.

```{r recipe}
rec <- recipe(y ~ x1 + x2, data = train) |>
  step_normalize(all_predictors())
```

## 3 — Model spec with `tune()`

`cost` is a `tune()` placeholder that maps to the $\Gamma$ regularisation
parameter inside `rmspe_lssvr()`.  The RBF bandwidth `rbf_sigma` is fixed at
1; it can also be tuned with `tune()`.

```{r spec}
spec <- psvr_rmspe_rbf(cost = tune(), rbf_sigma = 1) |>
  set_engine("psvr")
```

## 4 — Workflow

```{r workflow}
wf <- workflow() |>
  add_recipe(rec) |>
  add_model(spec)
```

## 5 — Tune with 5-fold CV

We search five candidate values of `cost` and evaluate each fold by MAPE.
`rmspe_lssvr()` only solves a linear system, so 25 fits complete in seconds.

```{r tune}
set.seed(2)
folds <- vfold_cv(train, v = 5)

cost_grid <- tibble::tibble(cost = c(0.01, 0.1, 1, 10, 100))

tune_res <- tune_grid(
  wf,
  resamples = folds,
  grid      = cost_grid,
  metrics   = metric_set(mape)
)
```

Cross-validated MAPE for each candidate (lower is better):

```{r tune-results}
collect_metrics(tune_res)[, c("cost", "mean", "std_err")]
```

## 6 — Select best

```{r best}
best_params <- select_best(tune_res, metric = "mape")
best_params
```

## 7 — Final fit and test-set evaluation

`last_fit()` refits on the full training set with the chosen `cost` and
evaluates once on the held-out test data.

```{r final}
final_wf  <- finalize_workflow(wf, best_params)
final_fit <- last_fit(final_wf, split, metrics = metric_set(mape))

collect_metrics(final_fit)
```

Predictions on the test set:

```{r predictions}
preds <- collect_predictions(final_fit)
head(preds[, c(".row", "y", ".pred")])
```

The fitted workflow can also be used directly for new data:

```{r new-data}
new_obs <- data.frame(x1 = c(0, 1, -2), x2 = c(0, 1, 2))
predict(extract_workflow(final_fit), new_data = new_obs)
```

## 8 — Inspecting the fitted psvr model

The tidymodels layer wraps a `psvr_rmspe` object (returned by `rmspe_lssvr()`). Extract it to use `print()` and `coef()` directly.

```{r engine-fit}
# extract_fit_engine() unwraps the parsnip/workflow layer to the raw psvr object
psvr_fit <- extract_fit_engine(extract_workflow(final_fit))
print(psvr_fit)
```

```{r engine-coef}
cf <- coef(psvr_fit)
# alpha: N dual variables; weight each training point in f(x) = sum_k alpha_k K(x_k, x) + b
# b:     bias / intercept term
# X_sv:  all N training inputs (LS-SVR has no sparsity — every training point contributes)
cat(sprintf("b = %.4f  |  alpha range: [%.4f, %.4f]\n",
            cf$b, min(cf$alpha), max(cf$alpha)))
```
