---
title: "Case Study: Boston Housing"
subtitle: "Benchmarking psvr models with percentage-error losses"
author: "Pablo Benavides-Herrera"
date: last-modified
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: kable
---

::: {.callout-note}
## Reproducibility note
These results correspond to the manuscript *"A Unified Family of
Percentage-Error Support Vector Regression Models with Symmetric
Kernel Extensions"* submitted to *Mathematics* (MDPI), currently
under review. Hyperparameter grids, random seeds, and model
implementations match those reported in the paper exactly.
Final published results may differ if revisions are requested.
Source code and raw results are available as downloadable files
at the end of this article.
:::

## 1 Introduction

The Boston Housing dataset (Harrison & Rubinfeld, 1978) is a canonical
regression benchmark containing 506 observations across 13 predictors.
The target variable, MEDV (median home value in $1,000s), is strictly
positive, making it a natural candidate for percentage-error evaluation.

Absolute-error metrics such as MSE treat a $2,000 prediction error the
same regardless of whether the true value is $10,000 or $100,000.
Percentage-error metrics — MAPE and RMSPE — penalize that same error
proportionally to the target scale, which better reflects real-estate
valuation decisions where relative accuracy drives business outcomes.

This article demonstrates that `psvr` models — trained directly under
MAPE and RMSPE loss — consistently outperform standard baselines on
those native percentage-error metrics over 30 randomized 80/20
train/test splits. The data are loaded from `mlbench::BostonHousing`.

## 2 Setup

```{r}
#| label: setup

library(tidyverse)
library(psvr)
library(mlbench)
library(e1071)
library(ranger)
library(xgboost)
library(quantreg)
library(knitr)
library(kableExtra)
library(xfun)
library(patchwork)

source("case-studies/experiment_helpers.R")

theme_set(theme_bw(base_size = 12))
```

## 3 Data

```{r}
#| label: data

data("BostonHousing", package = "mlbench")

df <- BostonHousing |>
  mutate(chas = as.integer(chas))

y <- df$medv
X <- df |> select(-medv) |> as.matrix()

stopifnot(all(y > 0))
```

```{r}
#| label: data-summary
#| echo: false

tibble::tibble(
  Statistic = c("n", "p", "min(y)", "median(y)", "mean(y)", "max(y)"),
  Value     = c(
    nrow(X), ncol(X),
    min(y), median(y), mean(y), max(y)
  )
) |>
  mutate(Value = round(Value, 2)) |>
  knitr::kable(caption = "Boston Housing dataset summary.")
```

```{r}
#| label: data-histogram
#| echo: false
#| fig-width: 7
#| fig-height: 4

ggplot(data.frame(y = y), aes(x = y)) +
  geom_histogram(binwidth = 2, fill = "#2c7bb6", color = "white",
                 alpha = 0.85) +
  labs(x = "MEDV ($1,000s)", y = "Count",
       title = "Distribution of Boston Housing target (MEDV)")
```

All 506 target values satisfy `y > 0` (confirmed by `stopifnot` above).

## 4 Experimental protocol

The benchmark follows a 30-seed protocol: for each seed in 1–30, the
506 observations are randomly split into 80% training (≈ 405 rows) and
20% test (≈ 101 rows). Hyperparameters are selected via 5-fold
cross-validation on the training set, minimizing each model's native CV
metric. Grid sizes are 48 combinations for SVR-MAPE, SVR-MAPE+SK,
ε-SVR (MSE), and Log-SVR; 16 for the four LS-RMSPE variants; up to 6
for Random Forest; 8 for XGBoost; and no tuning for Linear Regression,
WLS, and Quantile Regression. Six metrics are recorded on the held-out
test set: MAPE, RMSPE, MAAPE, MASE, MSE, and R². Prior to fitting,
predictor features are standardized to zero mean and unit variance
using statistics computed on the training fold only; test-set features
are scaled with the same parameters to prevent data leakage. MASE uses
the lag-1 naive denominator; because the Boston Housing data has no
temporal ordering, the denominator depends on row order within each
split and should be interpreted accordingly.

## 5 Run experiment

```{r}
#| label: experiment

results_file <- "case-studies/results/boston-results.csv"

if (!file.exists(results_file)) {
  results <- run_experiment(X, y,
                            dataset_name = "boston",
                            seeds        = 1:30,
                            verbose      = TRUE)
  write_csv(results, results_file)
} else {
  results <- read_csv(results_file, show_col_types = FALSE)
}

cat("Rows loaded:", nrow(results), "\n")
cat("Expected:   ", 30 * 13, "\n")
cat("NA count:   ", sum(is.na(results$MAPE)), "\n")
```

## 6 Results

### 6.1 Summary table

```{r}
#| label: summary-table

summary_df <- summarise_results(results)
```

```{r}
#| label: summary-table-display
#| echo: false

options(knitr.kable.NA = "—")

fmt_ci <- function(mn, lo, hi) {
  sprintf("%.2f [%.2f, %.2f]", mn, lo, hi)
}

bold_min <- function(x) {
  idx <- which.min(as.numeric(sub(" .*", "", x)))
  x[idx] <- cell_spec(x[idx], bold = TRUE)
  x
}

bold_max <- function(x) {
  idx <- which.max(as.numeric(x))
  x[idx] <- cell_spec(x[idx], bold = TRUE)
  x
}

tbl <- summary_df |>
  transmute(
    Model = label,
    MAPE  = fmt_ci(MAPE_mean,  MAPE_lower,  MAPE_upper),
    RMSPE = fmt_ci(RMSPE_mean, RMSPE_lower, RMSPE_upper),
    MAAPE = fmt_ci(MAAPE_mean, MAAPE_lower, MAAPE_upper),
    MSE   = sprintf("%.2f",  MSE_mean),
    R2    = sprintf("%.4f",  R2_mean)
  ) |>
  mutate(
    MAPE  = bold_min(MAPE),
    RMSPE = bold_min(RMSPE),
    MAAPE = bold_min(MAAPE),
    MSE   = bold_min(MSE),
    R2    = bold_max(R2)
  )

knitr::kable(
  tbl,
  format  = "html",
  escape  = FALSE,
  caption = paste(
    "Values in brackets are 95% percentile bootstrap confidence",
    "intervals over 30 random train/test splits."
  )
) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            full_width = FALSE)
```

### 6.2 Box plots

```{r}
#| label: boxplots
#| fig-width: 12
#| fig-height: 8

make_bp <- function(metric, ylab) {
  results |>
    select(abbrev, family, value = all_of(metric)) |>
    mutate(abbrev = fct_reorder(abbrev, value, .fun = median)) |>
    ggplot(aes(x = abbrev, y = value, fill = family)) +
    geom_boxplot(outlier.size = 0.8, linewidth = 0.4) +
    scale_fill_manual(values = c("psvr" = "#2c7bb6",
                                 "baseline" = "#d7191c"),
                      name = NULL) +
    labs(x = NULL, y = ylab, title = metric) +
    theme_bw(base_size = 11) +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
}

p_mape  <- make_bp("MAPE",  "MAPE (%)")
p_rmspe <- make_bp("RMSPE", "RMSPE (%)")
p_maape <- make_bp("MAAPE", "MAAPE (%)")
p_r2    <- make_bp("R2",    "R²")

(p_mape + p_rmspe) / (p_maape + p_r2) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### 6.3 Statistical comparison

```{r}
#| label: wilcoxon

wilcox_df <- wilcoxon_vs_best(results, metric = "MAPE")
```

```{r}
#| label: wilcoxon-table
#| echo: false

wilcox_df |>
  left_join(
    MODEL_LABELS |> select(id, ref_label = label),
    by = c("reference" = "id")
  ) |>
  select(
    Model                = label,
    `Reference baseline` = ref_label,
    `p-value`            = p_value,
    Significance         = signif
  ) |>
  mutate(`p-value` = sprintf("%.4f", `p-value`)) |>
  knitr::kable(
    caption = paste("Paired Wilcoxon signed-rank test vs.",
                    "best baseline on MAPE.")
  )
```

The best baseline in this study is
`r MODEL_LABELS$label[MODEL_LABELS$id == wilcox_df$reference[1]]`.
Of the six `psvr` models,
`r sum(wilcox_df$model_id %in% c("m1","m2","m3a","m3b","m4a","m4b") & wilcox_df$p_value < 0.05, na.rm = TRUE)`
achieve a statistically significant improvement over that baseline at
α = 0.05. Significance codes: \*\*\* p < 0.001, \*\* p < 0.01,
\* p < 0.05, ns = not significant.

## 7 Discussion

SVR-MAPE consistently achieves the lowest MAPE across splits because
its dual objective directly minimizes a weighted absolute-percentage
loss; no post-hoc transformation is required. LS-RMSPE with MAPE-based
CV tuning (LS-RMSPE-M) ranks closely behind, benefiting from the
closed-form linear system that avoids QP solver sensitivity. Models
with symmetric kernels — SVR-MAPE+SK and LS-RMSPE+SK — tend to show
tighter bootstrap confidence intervals, suggesting reduced variance
across the 30 splits even when their mean performance is comparable.

As expected, R² is slightly lower for `psvr` models than for
XGBoost or Random Forest: optimizing MAPE trades some explained
variance for scale-free accuracy. This is a deliberate design choice
and not a flaw — when MAPE is the deployment metric (e.g., property
valuation models reported to clients in percentage terms), `psvr`
models are the appropriate choice. For applications where R² or MSE
drives decisions, ensemble methods remain competitive.

**Practical recommendation:** use SVR-MAPE or LS-RMSPE when MAPE is
the target deployment metric; prefer XGBoost or Random Forest when R²
matters most.

## 8 Downloadable files

```{r}
#| label: downloads

xfun::embed_file(
  "case-studies/experiment_helpers.R",
  text = "Download experiment helper script (.R)"
)

xfun::embed_file(
  "case-studies/results/boston-results.csv",
  text = "Download full results (30 seeds x 13 models x 6 metrics)"
)
```
