---
title: "Boston Housing — reflection-augmented 2x2 ablation"
subtitle: "Validating Theorems 2 & 4 (kernel-symmetry vs. data-augmentation)"
author: "Pablo Benavides-Herrera"
date: last-modified
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: kable
---

::: {.callout-note}
## Reproducibility note
This article reproduces the 2x2 ablation experiment of Section 4.4 of the
manuscript *"A Unified Family of Percentage-Error Support Vector Regression
Models with Symmetric Kernel Extensions"* (under review at MDPI Mathematics).
It validates Theorems 2 and 4 by isolating two contributions to test-set
accuracy: the kernel-level symmetry constraint vs. the data-level reflection
augmentation. See the `boston_ablation_helpers.R` source for implementation
details.
:::

## 1 Introduction

The symmetric-kernel extensions of Theorems 2 (epsilon-SVR-MAPE) and 4
(LS-SVR-RMSPE) impose `f(x) = f(-x)` (the `a = +1` case) by replacing the
RBF kernel with the symmetrized kernel `K_s(x_i, x_j) = K(x_i, x_j) +
K(x_i, -x_j)`. A natural alternative is to enforce the same symmetry by
*augmenting the training set* with reflected points `(-x_k, y_k)` and
fitting a non-symmetric model on the doubled data. The two approaches
agree at the dual level (Theorem 5), but they differ in computational and
statistical behavior. This experiment empirically separates the two.

For each of two model families — A: epsilon-SVR-MAPE; B: LS-SVR-RMSPE — we
evaluate four cells in a 2x2 design over symmetry (kernel-level) and
augmentation (data-level):

|              | non-augmented (size N) | augmented (size 2N)             |
|--------------|------------------------|----------------------------------|
| non-symmetric| **cell 1** (baseline)  | **cell 2** (data-aug only)       |
| symmetric    | **cell 3** (kernel only) | **cell 4** (joint)             |

Eight cells total (A1...A4, B1...B4). All cells use the same RBF kernel
family, the same outer 80/20 split per seed, the same `cost` and
`rbf_sigma` search ranges within a family, and Bayesian-optimization-style
tuning with a budget of 50 inner-CV evaluations per cell.

The key empirical question: does the kernel symmetry constraint contribute
more (Δ_sym), the data augmentation contribute more (Δ_aug), or are they
roughly equivalent (Δ_aug ≈ Δ_sym)? Theorem 5 predicts the former.

## 2 Setup

```{r}
#| label: setup

library(tidyverse)
library(tidymodels)
library(psvr)
library(mlbench)
library(furrr)
library(future)
library(here)
library(knitr)
library(kableExtra)

here::i_am("vignettes/articles/boston-reflection-ablation.qmd")
source(here::here("vignettes/articles/case-studies/experiment_helpers.R"))
source(here::here("vignettes/articles/case-studies/boston_ablation_helpers.R"))

tidymodels_prefer()
theme_set(theme_bw(base_size = 12))
```

## 3 Configuration

```{r}
#| label: config

N_SEEDS   <- 30L          # set to 1L for sanity check (sequential)
N_WORKERS <- 12L          # match run_all_experiments.R
EXCLUDE_B <- FALSE        # match boston-housing.qmd
LHS_BUDGET <- 50L         # per-cell HP evaluations for augmented cells
```

## 4 Data

```{r}
#| label: data

data("BostonHousing", package = "mlbench")
df <- BostonHousing |> mutate(chas = as.integer(chas))
if (EXCLUDE_B) {
  message("Dropping column B (EXCLUDE_B = TRUE).")
  df <- df |> select(-b)
}
y <- df$medv
X <- df |> select(-medv) |> as.matrix()
stopifnot(all(y > 0))
```

```{r}
#| label: data-summary
#| echo: false

tibble::tibble(
  Statistic = c("n", "p", "min(y)", "median(y)", "mean(y)", "max(y)"),
  Value     = c(nrow(X), ncol(X), min(y), median(y), mean(y), max(y))
) |>
  mutate(Value = round(Value, 2)) |>
  knitr::kable(caption = "Boston Housing: dataset summary (post-EXCLUDE_B).")
```

## 5 Methodology

### 5.1 The 2x2 ablation cells

Each of the eight cells fits a single psvr model under one combination of
(symmetric kernel) and (reflection augmentation):

|cell | family            | model fn                  | sym (a) | augmented training data         |
|-----|-------------------|---------------------------|---------|----------------------------------|
| A1  | epsilon-SVR-MAPE  | `psvr::mape_svr`          | none    | `(X_tr, y_tr)`                  |
| A2  | epsilon-SVR-MAPE  | `psvr::mape_svr`          | none    | `(rbind(X_tr, -X_tr), c(y_tr, y_tr))` |
| A3  | epsilon-SVR-MAPE  | `psvr::mape_sym_svr`      | +1      | `(X_tr, y_tr)`                  |
| A4  | epsilon-SVR-MAPE  | `psvr::mape_sym_svr`      | +1      | `(rbind(X_tr, -X_tr), c(y_tr, y_tr))` |
| B1  | LS-SVR-RMSPE      | `psvr::rmspe_lssvr`       | none    | `(X_tr, y_tr)`                  |
| B2  | LS-SVR-RMSPE      | `psvr::rmspe_lssvr`       | none    | `(rbind(X_tr, -X_tr), c(y_tr, y_tr))` |
| B3  | LS-SVR-RMSPE      | `psvr::rmspe_sym_lssvr`   | +1      | `(X_tr, y_tr)`                  |
| B4  | LS-SVR-RMSPE      | `psvr::rmspe_sym_lssvr`   | +1      | `(rbind(X_tr, -X_tr), c(y_tr, y_tr))` |

Symmetry type `a = +1` ("even") is fixed in cells 3 & 4 (not tuned).

### 5.2 Standardization

`recipes::step_normalize` convention (denominator N-1) on the
**non-augmented** training fold, then augmentation in standardized space:

  1. compute `(μ, σ)` from the non-augmented inner-training rows
  2. transform inner-validation rows with the same `(μ, σ)`
  3. for augmented cells: `Z = (X_tr - μ) / σ`, then fit on `rbind(Z, -Z)`

This guarantees that all 8 cells share the same `(μ, σ)` source per seed —
only the {Z, -Z} duplication differs.

### 5.3 Hyperparameter tuning

| family | search range (transformed)                          | tuner          | budget       |
|--------|-----------------------------------------------------|----------------|--------------|
| A      | `cost`: log2 [-2, 10] (psvr::cost_psvr)             | tune_bayes (cells A1, A3) / 50-pt LHS (cells A2, A4) | 15+35 BO / 50 LHS |
|        | `rbf_sigma`: log10 from `psvr::rbf_sigma_psvr_data` |                |              |
|        | `svm_margin`: linear [1, 20]% (margin_percentage)   |                |              |
| B      | `cost`: log2 [-2, log2(var(y_tr) * N) + 4]          | tune_bayes (cells B1, B3) / 50-pt LHS (cells B2, B4) | 10+40 BO / 50 LHS |
|        | `rbf_sigma`: same as A                              |                |              |

Selection metric for all 8 cells: MAPE.

### 5.4 Why two different tuners?

Augmented cells (A2, A4, B2, B4) require **paired 5-fold CV** to avoid
inner-CV leakage: with naive 5-fold over the 2N augmented rows, mirrored
pairs `(x_i, -x_i)` can split across folds, putting `x_i` in inner-train
and `-x_i` in inner-val — the symmetric-kernel model has effectively
seen `y_i`. Paired CV constructs the 5 folds on the original N rows and
augments only the inner-training portion within each fold. `tune_bayes`
does not support row-multiplying training-only recipe steps, so paired CV
is hand-rolled with a 50-point Latin hypercube over the same
parameter ranges as the non-augmented cells.

::: {.callout-warning}
## Methodological caveat
LHS-50 ≠ tune_bayes(initial=15, iter=35). At this budget, `tune_bayes`'
GP+EI surrogate typically reaches lower validation MAPE than uniform
LHS sampling. This biases the comparison **against** augmented cells
(better tuning → lower MAPE in the non-augmented arm). If Δ_sym
nonetheless dominates Δ_aug despite this anti-symmetric bias, the
conclusion is robust.
:::

### 5.5 Test-leak assertion

After standardization and augmentation, an explicit `stopifnot` checks
that no row of the (possibly reflected) training fold appears in the
test fold. With continuous Boston features this assertion's purpose is
to catch indexing bugs.

## 6 Run experiment

```{r}
#| label: run-experiment

results_file <- here::here(
  "vignettes/articles/case-studies/results",
  "boston-reflection-ablation-results.csv"
)

if (!file.exists(results_file)) {
  results <- run_experiment_ablation_parallel(
    X, y,
    seeds        = seq_len(N_SEEDS),
    workers      = N_WORKERS,
    dataset_name = "boston_ablation"
  )
} else {
  results <- read_csv(results_file, show_col_types = FALSE)
}

cat("Rows loaded:", nrow(results), "  (expected:", 8L * N_SEEDS, ")\n")
cat("NA-MAPE rows:", sum(is.na(results$MAPE)), "\n")
```

## 7 Per-cell summary table

```{r}
#| label: cell-summary

cell_summary <- results |>
  group_by(family, cell_id, sym, augmented) |>
  summarise(
    MAPE_mean  = mean(MAPE, na.rm = TRUE),
    MAPE_lower = bootstrap_ci(MAPE)[["lower"]],
    MAPE_upper = bootstrap_ci(MAPE)[["upper"]],
    R2_mean    = mean(R2,   na.rm = TRUE),
    .groups    = "drop"
  ) |>
  arrange(family, cell_id)

cell_summary |>
  mutate(
    MAPE = sprintf("%.2f [%.2f, %.2f]", MAPE_mean, MAPE_lower, MAPE_upper),
    R2   = sprintf("%.4f", R2_mean)
  ) |>
  select(family, cell_id, sym, augmented, MAPE, R2) |>
  knitr::kable(
    caption = sprintf(
      "Per-cell mean MAPE with 95%% percentile bootstrap CI over %d seeds.",
      N_SEEDS
    )
  )
```

## 8 Δ-decomposition (Δ_aug, Δ_sym, Δ_joint) per family

For each family, compute the per-seed differences relative to cell 1
(non-augmented, non-symmetric) and report mean ± bootstrap CI.

```{r}
#| label: deltas

delta_decomposition <- function(results, family_letter) {
  fam <- results |>
    filter(family == family_letter) |>
    select(seed, cell_id, MAPE) |>
    pivot_wider(names_from = cell_id, values_from = MAPE)

  cells <- if (family_letter == "A") c("A1", "A2", "A3", "A4")
           else                       c("B1", "B2", "B3", "B4")
  d_aug   <- fam[[cells[1]]] - fam[[cells[2]]]
  d_sym   <- fam[[cells[1]]] - fam[[cells[3]]]
  d_joint <- fam[[cells[1]]] - fam[[cells[4]]]

  fmt <- function(x) {
    bs <- bootstrap_ci(x)
    sprintf("%+.3f [%+.3f, %+.3f]", bs[["mean"]], bs[["lower"]], bs[["upper"]])
  }

  tibble::tibble(
    family   = family_letter,
    delta    = c("Δ_aug (cell1 − cell2)",
                 "Δ_sym (cell1 − cell3)",
                 "Δ_joint (cell1 − cell4)"),
    estimate = c(fmt(d_aug), fmt(d_sym), fmt(d_joint))
  )
}

bind_rows(
  delta_decomposition(results, "A"),
  delta_decomposition(results, "B")
) |>
  knitr::kable(
    caption = "Δ-decomposition (positive = improvement over cell 1 baseline)."
  )
```

## 9 Paired Wilcoxon vs. cell-1 baseline

```{r}
#| label: wilcoxon

wilcoxon_per_family <- function(results, family_letter) {
  fam <- results |>
    filter(family == family_letter) |>
    select(seed, cell_id, MAPE) |>
    pivot_wider(names_from = cell_id, values_from = MAPE)

  baseline <- if (family_letter == "A") "A1" else "B1"
  comparisons <- if (family_letter == "A") c("A2", "A3", "A4")
                 else                       c("B2", "B3", "B4")

  do.call(rbind, lapply(comparisons, function(c2) {
    test <- stats::wilcox.test(fam[[baseline]], fam[[c2]],
                               paired = TRUE, exact = FALSE)
    tibble::tibble(
      family    = family_letter,
      cell      = c2,
      reference = baseline,
      mean_diff = mean(fam[[baseline]] - fam[[c2]], na.rm = TRUE),
      p_value   = test$p.value,
      signif    = case_when(
        test$p.value < 0.001 ~ "***",
        test$p.value < 0.01  ~ "**",
        test$p.value < 0.05  ~ "*",
        TRUE                 ~ "ns"
      )
    )
  }))
}

bind_rows(
  wilcoxon_per_family(results, "A"),
  wilcoxon_per_family(results, "B")
) |>
  mutate(
    mean_diff = sprintf("%+.3f", mean_diff),
    p_value   = sprintf("%.4f",  p_value)
  ) |>
  knitr::kable(
    caption = "Paired Wilcoxon signed-rank test, each non-baseline cell vs cell 1."
  )
```

## 10 Discussion

The Δ-decomposition table is the empirical answer to the question
"does the symmetry come from the kernel or the data?":

- **Δ_sym dominates Δ_aug** → kernel constraint is the load-bearing
  contribution. This is what Theorem 5 predicts.
- **Δ_aug dominates Δ_sym** → data augmentation alone is sufficient;
  the kernel-level constraint adds little. Theorem 5 would still hold
  formally (the dual problems are equivalent at the optimum), but the
  finite-sample / finite-budget statistical behavior would favor
  augmentation as the practical lever.
- **Δ_aug ≈ Δ_sym ≈ Δ_joint / 2** → the two effects are additive and
  roughly equal. Both are useful; neither dominates.

Δ_joint should be approximately Δ_sym + Δ_aug minus an interaction term.
Negative interaction (Δ_joint < Δ_sym + Δ_aug) is expected: applying the
symmetric kernel to already-augmented data is partially redundant, since
the symmetric kernel already encodes the same `f(x) = f(-x)` constraint
that the augmentation enforces statistically.

## 11 Downloadable files

```{r}
#| label: downloads
#| eval: !expr file.exists(results_file)

xfun::embed_file(
  here::here("vignettes/articles/case-studies/boston_ablation_helpers.R"),
  text = "Download cell-runner helpers (.R)"
)

xfun::embed_file(
  results_file,
  text = sprintf("Download full results (%d seeds × 8 cells)", N_SEEDS)
)
```
