---
title: "Case Study: Diabetes Progression"
subtitle: "Benchmarking psvr models with percentage-error losses"
author: "Pablo Benavides-Herrera"
date: last-modified
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: kable
---

::: {.callout-note}
## Reproducibility note
These results correspond to the manuscript *"A Unified Family of
Percentage-Error Support Vector Regression Models with Symmetric
Kernel Extensions"* submitted to *Mathematics* (MDPI), currently
under review. Hyperparameter grids, random seeds, and model
implementations match those reported in the paper exactly.
Final published results may differ if revisions are requested.
Source code and raw results are available as downloadable files
at the end of this article.
:::

## 1 Introduction

The Diabetes dataset (Efron et al., 2004) comprises n = 442
patients with p = 10 baseline clinical measurements (age, sex,
BMI, blood pressure, and six serum measurements). The target is
a quantitative measure of disease progression one year after
baseline, which is strictly positive with high relative
variability (CV ≈ 51%). In clinical decision support, relative
prediction error is more meaningful than absolute error: a
10-point error for a low-progression patient carries different
clinical weight than the same error for a high-progression
patient.

This article demonstrates that `psvr` models — trained directly under
MAPE and RMSPE loss — provide competitive percentage-error performance
over 30 randomized train/test splits on this challenging small-sample
regression problem. Data loaded from `lars::diabetes`.

## 2 Setup

```{r}
#| label: setup

library(tidyverse)
library(psvr)
library(lars)
library(e1071)
library(ranger)
library(xgboost)
library(quantreg)
library(knitr)
library(kableExtra)
library(xfun)
library(patchwork)

source("case-studies/experiment_helpers.R")

theme_set(theme_bw(base_size = 12))
```

## 3 Data

```{r}
#| label: data

data("diabetes", package = "lars")
y <- as.numeric(diabetes$y)
X <- as.matrix(diabetes$x)

stopifnot(all(y > 0))
```

```{r}
#| label: data-summary
#| echo: false

tibble::tibble(
  Statistic = c("n", "p", "min(y)", "median(y)", "mean(y)", "max(y)"),
  Value     = c(
    nrow(X), ncol(X),
    min(y), median(y), mean(y), max(y)
  )
) |>
  mutate(Value = round(Value, 2)) |>
  knitr::kable(caption = "Diabetes dataset summary.")
```

```{r}
#| label: data-histogram
#| echo: false
#| fig-width: 7
#| fig-height: 4

ggplot(data.frame(y = y), aes(x = y)) +
  geom_histogram(binwidth = 20, fill = "#2c7bb6", color = "white",
                 alpha = 0.85) +
  labs(x = "Disease Progression", y = "Count",
       title = "Distribution of Diabetes target (disease progression)")
```

All 442 target values satisfy `y > 0` (confirmed by `stopifnot` above).

::: {.callout-note}
The coefficient of variation of the target (CV ≈ 51%) is substantially
higher than Boston Housing, making percentage-error metrics more
challenging but also more informative for this dataset.
:::

## 4 Experimental protocol

The benchmark follows a 30-seed protocol: for each seed in 1–30, the
442 observations are randomly split into 80% training (≈ 353 rows) and
20% test (≈ 89 rows). Hyperparameters are selected via 5-fold
cross-validation on the training set, minimizing each model's native CV
metric. Grid sizes are 48 combinations for SVR-MAPE, SVR-MAPE+SK,
ε-SVR (MSE), and Log-SVR; 16 for the four LS-RMSPE variants; up to 6
for Random Forest; 8 for XGBoost; and no tuning for Linear Regression,
WLS, and Quantile Regression. Six metrics are recorded on the held-out
test set: MAPE, RMSPE, MAAPE, MASE, MSE, and R². Prior to fitting,
predictor features are standardized to zero mean and unit variance
using statistics computed on the training fold only; test-set features
are scaled with the same parameters to prevent data leakage. MASE uses
the lag-1 naive denominator; because the Diabetes data has no temporal
ordering, the denominator depends on row order within each split and
should be interpreted accordingly.

## 5 Run experiment

```{r}
#| label: experiment

results_file <- "case-studies/results/diabetes-results.csv"

if (!file.exists(results_file)) {
  results <- run_experiment(X, y,
                            dataset_name = "diabetes",
                            seeds        = 1:30,
                            verbose      = TRUE)
  write_csv(results, results_file)
} else {
  results <- read_csv(results_file, show_col_types = FALSE)
}

cat("Rows loaded:", nrow(results), "\n")
cat("Expected:   ", 30 * 13, "\n")
cat("NA count:   ", sum(is.na(results$MAPE)), "\n")
```

## 6 Results

### 6.1 Summary table

```{r}
#| label: summary-table

summary_df <- summarise_results(results)
```

```{r}
#| label: summary-table-display
#| echo: false

options(knitr.kable.NA = "—")

fmt_ci <- function(mn, lo, hi) {
  sprintf("%.2f [%.2f, %.2f]", mn, lo, hi)
}

bold_min <- function(x) {
  idx <- which.min(as.numeric(sub(" .*", "", x)))
  x[idx] <- cell_spec(x[idx], bold = TRUE)
  x
}

bold_max <- function(x) {
  idx <- which.max(as.numeric(x))
  x[idx] <- cell_spec(x[idx], bold = TRUE)
  x
}

tbl <- summary_df |>
  transmute(
    Model = label,
    MAPE  = fmt_ci(MAPE_mean,  MAPE_lower,  MAPE_upper),
    RMSPE = fmt_ci(RMSPE_mean, RMSPE_lower, RMSPE_upper),
    MAAPE = fmt_ci(MAAPE_mean, MAAPE_lower, MAAPE_upper),
    MSE   = sprintf("%.2f",  MSE_mean),
    R2    = sprintf("%.4f",  R2_mean)
  ) |>
  mutate(
    MAPE  = bold_min(MAPE),
    RMSPE = bold_min(RMSPE),
    MAAPE = bold_min(MAAPE),
    MSE   = bold_min(MSE),
    R2    = bold_max(R2)
  )

knitr::kable(
  tbl,
  format  = "html",
  escape  = FALSE,
  caption = paste(
    "Values in brackets are 95% percentile bootstrap confidence",
    "intervals over 30 random train/test splits."
  )
) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            full_width = FALSE)
```

### 6.2 Box plots

```{r}
#| label: boxplots
#| fig-width: 12
#| fig-height: 8

make_bp <- function(metric, ylab) {
  results |>
    select(abbrev, family, value = all_of(metric)) |>
    mutate(abbrev = fct_reorder(abbrev, value, .fun = median)) |>
    ggplot(aes(x = abbrev, y = value, fill = family)) +
    geom_boxplot(outlier.size = 0.8, linewidth = 0.4) +
    scale_fill_manual(values = c("psvr" = "#2c7bb6",
                                 "baseline" = "#d7191c"),
                      name = NULL) +
    labs(x = NULL, y = ylab, title = metric) +
    theme_bw(base_size = 11) +
    theme(axis.text.x  = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
}

p_mape  <- make_bp("MAPE",  "MAPE (%)")
p_rmspe <- make_bp("RMSPE", "RMSPE (%)")
p_maape <- make_bp("MAAPE", "MAAPE (%)")
p_r2    <- make_bp("R2",    "R²")

(p_mape + p_rmspe) / (p_maape + p_r2) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### 6.3 Statistical comparison

```{r}
#| label: wilcoxon

wilcox_df <- wilcoxon_vs_best(results, metric = "MAPE")
```

```{r}
#| label: wilcoxon-table
#| echo: false

wilcox_df |>
  left_join(
    MODEL_LABELS |> select(id, ref_label = label),
    by = c("reference" = "id")
  ) |>
  select(
    Model                = label,
    `Reference baseline` = ref_label,
    `p-value`            = p_value,
    Significance         = signif
  ) |>
  mutate(`p-value` = sprintf("%.4f", `p-value`)) |>
  knitr::kable(
    caption = paste("Paired Wilcoxon signed-rank test vs.",
                    "best baseline on MAPE.")
  )
```

The best baseline in this study is
`r MODEL_LABELS$label[MODEL_LABELS$id == wilcox_df$reference[1]]`.
Of the six `psvr` models,
`r sum(wilcox_df$model_id %in% c("m1","m2","m3a","m3b","m4a","m4b") & wilcox_df$p_value < 0.05, na.rm = TRUE)`
achieve a statistically significant improvement over that baseline at
α = 0.05. Significance codes: \*\*\* p < 0.001, \*\* p < 0.01,
\* p < 0.05, ns = not significant.

## 7 Discussion

SVR-MAPE leads on MAPE (34–35% range) despite the dataset's high CV;
predicting always the mean yields approximately 62% MAPE, so `psvr`
models capture meaningful signal. R² is modest (0.3–0.5) for all
models — expected for this small, noisy dataset where linear methods
are competitive.

Symmetric kernel models (m2, m4a, m4b) show R² near zero or negative:
Diabetes has no natural symmetry structure, so the symmetric kernel
term introduces noise rather than useful inductive bias. This delimits
the applicability of symmetric extensions to datasets with inherent
symmetry in the feature space.

As with Boston Housing, optimizing MAPE trades some explained variance
for scale-free accuracy. SVR-MAPE is the appropriate model when
percentage-error is the clinical reporting metric. For R²-focused
tasks, linear regression remains competitive on this small dataset and
should be considered a strong baseline.

**Practical recommendation:** use SVR-MAPE when percentage-error is
the clinical reporting metric; for R²-focused tasks, linear regression
remains competitive on this dataset.

## 8 Downloadable files

```{r}
#| label: downloads

xfun::embed_file(
  "case-studies/experiment_helpers.R",
  text = "Download experiment helper script (.R)"
)

xfun::embed_file(
  "case-studies/results/diabetes-results.csv",
  text = "Download full results (30 seeds x 13 models x 6 metrics)"
)
```
