---
title: "Electricity Demand Forecasting with Percentage-Error SVR"
subtitle: "Applying psvr models to Victoria, Australia daily demand"
author: "Pablo Benavides-Herrera"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    code-fold: true
    code-tools: true
    code-link: true
    code-annotations: below
    theme: cosmo
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
    embed-resources: true
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    fig.width: 7
    fig.asp: 0.618
    out.width: "90%"
---

```{r}
#| label: setup
#| code-fold: true
library(tsibbledata)
library(tsibble)
library(lubridate)
library(dplyr)
library(ggplot2)
library(plotly)
library(knitr)
library(kableExtra)
library(kernlab)
library(scales)
library(psvr)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold"),
      plot.subtitle    = element_text(colour = "grey40"),
      panel.grid.minor = element_blank()
    )
)

vpal <- viridis_pal(option = "D")(6)

mape_fn  <- function(y, yhat) mean(abs(y - yhat) / y) * 100
rmspe_fn <- function(y, yhat) sqrt(mean(((y - yhat) / y)^2)) * 100
r2_fn    <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)
mse_fn   <- function(y, yhat) mean((y - yhat)^2)
```

---

## 1  The forecasting problem

Electricity grid operators must predict demand days ahead to schedule generation
assets, balance the grid, and avoid costly emergency purchases. Victoria,
Australia operates a competitive spot market where forecast errors translate
directly into monetary penalties — making **relative accuracy** the natural
objective.

The `vic_elec` dataset from the **tsibbledata** package records half-hourly
demand and temperature readings for the state of Victoria over 2012–2014. We
aggregate to the **daily** level: total daily demand (MWh) and mean daily
temperature (°C). Daily forecasting is a common planning horizon for unit
commitment and fuel scheduling.

```{r}
#| label: data-prep
#| code-fold: false

elec_daily <- vic_elec |>
  as_tibble() |>
  mutate(Date = as.Date(Time)) |>
  group_by(Date) |>
  summarise(
    Demand      = sum(Demand,       na.rm = TRUE),   # <1>
    Temperature = mean(Temperature, na.rm = TRUE),   # <2>
    IsHoliday   = any(Holiday),                      # <3>
    .groups = "drop"
  ) |>
  mutate(
    Temperature2 = Temperature^2,                    # <4>
    DayOfWeek    = wday(Date, week_start = 1),       # <5>
    Month        = month(Date),
    IsWeekend    = as.integer(DayOfWeek >= 6),
    IsHoliday    = as.integer(IsHoliday)
  )

cat(sprintf("Rows: %d  |  Dates: %s – %s\n",
            nrow(elec_daily), min(elec_daily$Date), max(elec_daily$Date)))
cat(sprintf("Demand range: %.0f – %.0f MWh/day\n",
            min(elec_daily$Demand), max(elec_daily$Demand)))
cat(sprintf("Public holidays: %d  |  Weekend days: %d\n",
            sum(elec_daily$IsHoliday), sum(elec_daily$IsWeekend)))
```

1. Sum all 48 half-hourly readings to get total daily demand in MWh.
2. Average temperature across the day.
3. `Holiday` in `vic_elec` is a logical vector — `any(Holiday)` is `TRUE` if
   any half-hour in that day is flagged as a public holiday.
4. Squared temperature to capture the U-shaped demand–temperature relationship
   (high AC demand in summer, high heating demand in winter).
5. `week_start = 1` → Monday = 1, Sunday = 7.

```{r}
#| label: fig-ts
#| fig-cap: "Daily electricity demand in Victoria, 2012–2014. Colour encodes mean daily temperature. The summer (Dec–Feb) heat-wave spikes and the lower weekend/holiday baseline are visible."
#| code-fold: true

p_ts <- elec_daily |>
  ggplot(aes(x = Date, y = Demand / 1e3, colour = Temperature,
             text = paste0("Date: ", Date,
                           "<br>Demand: ", round(Demand / 1e3, 1), " GWh",
                           "<br>Temp: ", round(Temperature, 1), "°C"))) +
  geom_line(linewidth = 0.4, alpha = 0.85) +
  scale_colour_viridis_c(option = "C", name = "Temp (°C)") +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Demand (GWh/day)",
       title = "Victoria daily electricity demand — 2012–2014",
       subtitle = "Colour = mean daily temperature") +
  theme_bw() +
  theme(legend.position = "right")

ggplotly(p_ts, tooltip = "text") |>
  layout(hoverlabel = list(bgcolor = "white"), width = NULL, height = 450)
```

::: {.callout-note}
## Why MAPE matters here

Absolute-error objectives (MAE, MSE) treat all days equally. But the grid
operator's cost function is relative:

- A **10,000 MWh** error on a mild autumn day (baseline ≈ 165,000 MWh) is
  **6.1% — critical**: the error is large relative to scheduled capacity.
- The **same 10,000 MWh** error on a hot summer day (baseline ≈ 305,000 MWh)
  is **3.3% — more tolerable**: dwarfed by total load.

Minimising MAPE directly aligns the model with the operator's penalty
structure.
:::

```{r}
#| label: fig-asymmetric-cost
#| fig-cap: "A fixed 10,000 MWh absolute error has very different relative impact depending on the day's baseline demand."
#| code-fold: true

tibble(
  baseline  = seq(80e3, 360e3, by = 1000),
  abs_error = 10000,
  pct_error = abs_error / baseline * 100
) |>
  ggplot(aes(baseline / 1e3, pct_error)) +
  geom_line(colour = vpal[2], linewidth = 1.2) +
  geom_point(
    data = tibble(
      baseline  = c(165e3, 305e3),
      pct_error = 10000 / c(165e3, 305e3) * 100
    ),
    aes(baseline / 1e3, pct_error), size = 3, colour = vpal[5]
  ) +
  annotate("text", x = 165, y = 10000 / 165e3 * 100 + 0.5,
           label = "Mild day: 6.1%", hjust = 0, size = 3.5) +
  annotate("text", x = 305, y = 10000 / 305e3 * 100 + 0.5,
           label = "Hot day: 3.3%",  hjust = 1, size = 3.5) +
  labs(x = "Baseline demand (GWh/day)", y = "Percentage error (%)",
       title = "Asymmetric impact of a fixed 10 GWh error",
       subtitle = "Same absolute error — very different relative cost") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_continuous(labels = comma)
```

---

## 2  Support Vector Regression: the intuition

Classical regression minimises residuals globally. SVR takes a different
approach: it finds the **flattest** function (smallest `‖ω‖²`) that fits all
training points **within an ε-tube**. Points outside the tube are penalised,
and only those points — the **support vectors** — determine the solution.

```{r}
#| label: fig-svr-intuition
#| fig-cap: "SVR on a 1-D synthetic dataset. Grey band: the ε-insensitivity tube. Blue points: support vectors (outside or on the boundary). The fitted line is determined only by the support vectors."
#| code-fold: true

set.seed(7)
n_syn  <- 40
x_syn  <- sort(runif(n_syn, 0, 10))
y_syn  <- 2 * x_syn + 1 + rnorm(n_syn, sd = 1.5)

df_syn   <- data.frame(x = x_syn, y = y_syn)
svm_1d   <- ksvm(y ~ x, data = df_syn, type = "eps-svr",
                 kernel = "vanilladot", C = 5, epsilon = 1.5, scaled = FALSE)
pred_syn <- predict(svm_1d, df_syn)
sv_flags <- seq_len(n_syn) %in% SVindex(svm_1d)

tibble(x = x_syn, y = y_syn, yhat = as.numeric(pred_syn), is_sv = sv_flags) |>
  ggplot(aes(x, y)) +
  geom_ribbon(aes(ymin = yhat - 1.5, ymax = yhat + 1.5),
              fill = "grey85", alpha = 0.6) +
  geom_line(aes(y = yhat), colour = vpal[2], linewidth = 1) +
  geom_point(aes(colour = is_sv, size = is_sv), alpha = 0.8) +
  scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = vpal[5]),
                      labels = c("Interior point", "Support vector"),
                      name = NULL) +
  scale_size_manual(values = c("FALSE" = 2, "TRUE" = 3.5), guide = "none") +
  labs(x = "x", y = "y",
       title = "SVR: fitting within the ε-tube",
       subtitle = "Only support vectors (blue) influence the decision boundary") +
  theme(legend.position = "top")
```

::: {.callout-tip collapse="true"}
## Mathematical detail: SVR primal formulation

The standard ε-SVR solves the **primal problem**:

$$
\min_{\omega, b, \xi, \xi^*} \quad \tfrac{1}{2}\|\omega\|^2
  + C \sum_{k=1}^N (\xi_k + \xi_k^*)
$$

subject to

$$
\begin{aligned}
y_k - \omega^\top \phi(x_k) - b &\leq \varepsilon + \xi_k \\
\omega^\top \phi(x_k) + b - y_k &\leq \varepsilon + \xi_k^* \\
\xi_k,\, \xi_k^* &\geq 0
\end{aligned}
$$

The ε-tube means residuals smaller than ε incur **zero loss** — only outliers
contribute to the objective. The dual reveals the solution
$f(x) = \sum_k \beta_k K(x_k, x) + b$ where $\beta_k = \alpha_k - \alpha_k^*$.
:::

---

## 3  Why percentage error changes the math

### Classical vs MAPE box constraints

In classical ε-SVR, the QP box constraints are **uniform**:
$$|\beta_k| \leq C \quad \text{for all } k.$$

In the MAPE ε-SVR (Model 1), the constraints become **per-sample**:
$$|\beta_k| \leq \frac{100\,C}{y_k}.$$

Low-demand days get tighter bounds — the model devotes more capacity to
accurately representing them, which is exactly what a percentage-error
objective requires. Note that all psvr models are fitted on **scaled targets**
(see Section 4); the figure below uses the scaled range.

```{r}
#| label: fig-box-constraints
#| fig-cap: "Per-sample box constraint for MAPE-SVR (100C/yₖ, with C = 5) versus the flat classical bound. At lower scaled demand (~1.6) the bound tightens substantially."
#| code-fold: true

C_val  <- 5
y_grid <- seq(1.5, 3.6, by = 0.01)   # scaled daily demand range

tibble(
  y         = y_grid,
  classical = C_val,
  mape_svr  = 100 * C_val / y_grid
) |>
  tidyr::pivot_longer(-y, names_to = "model", values_to = "bound") |>
  mutate(model = recode(model,
    classical = "Classical: |β| ≤ C",
    mape_svr  = "MAPE-SVR: |β| ≤ 100C/y"
  )) |>
  ggplot(aes(y, bound, colour = model, linetype = model)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c(vpal[2], vpal[5]), name = NULL) +
  scale_linetype_manual(values = c("dashed", "solid"), name = NULL) +
  labs(x = "Scaled target yₖ  (Demand / 1e5)", y = "Box constraint bound",
       title = "Box constraints: classical vs MAPE-SVR  (C = 5)",
       subtitle = "MAPE-SVR tightens constraints for lower-demand days") +
  theme(legend.position = "top")
```

::: {.callout-tip collapse="true"}
## Mathematical detail: MAPE ε-SVR dual (Theorem 1)

Replacing the absolute-error tube with a **percentage tube** —
$(100/y_k)|y_k - f(x_k)| \leq \varepsilon$ — transforms the dual to:

$$
\max_\beta \quad -\tfrac{1}{2} \sum_{k,l} \beta_k \beta_l K(x_k, x_l)
  + \sum_k \beta_k y_k
  - \frac{\varepsilon}{100} \sum_k |\beta_k|\, y_k
$$

subject to $\sum_k \beta_k = 0$ and $|\beta_k| \leq 100C/y_k$.

The per-sample upper bound $100C/y_k$ tightens for small $y_k$ (low-demand
days), preventing the model from over-fitting absolute fluctuations on those
days. See Theorem 1 of Benavides-Herrera et al. (2026) for the full derivation.
:::

### LS-SVR and the RMSPE loss

The LS-SVR variant (Model 3) replaces the ε-insensitive loss with a squared
loss, turning the QP into a **linear system**. The percentage-error variant
modifies the regularization diagonal:

$$Y_\Gamma = \text{diag}\!\left(\frac{y_1^2}{\Gamma}, \ldots, \frac{y_N^2}{\Gamma}\right)$$

This target-weighted diagonal causes the system to minimise the root mean
**squared** percentage error (RMSPE) rather than the standard squared residual.
The key advantage: the solution requires only `base::solve()` on an
$(N+1)\times(N+1)$ system — no QP solver needed.

---

## 4  Feature engineering and train/test split

Six features describe each day: temperature (linear and quadratic), day of
week, month, weekend flag, and public holiday flag. We use a strict temporal
split — all of 2012–2013 for training, all of 2014 for testing — to avoid
any look-ahead bias.

```{r}
#| label: split
#| code-fold: false

feature_cols <- c("Temperature", "Temperature2", "DayOfWeek",
                  "Month", "IsWeekend", "IsHoliday")

train <- elec_daily |> filter(year(Date) <= 2013)   # <1>
test  <- elec_daily |> filter(year(Date) == 2014)

X_raw_tr <- as.matrix(train[, feature_cols])
y_tr     <- train$Demand
X_raw_te <- as.matrix(test[, feature_cols])
y_te     <- test$Demand

# Standardise features using training statistics only           # <2>
cm   <- colMeans(X_raw_tr)
cs   <- apply(X_raw_tr, 2, sd)
X_tr <- scale(X_raw_tr, center = cm, scale = cs)
X_te <- scale(X_raw_te, center = cm, scale = cs)

# Scale targets to ~ [1.6, 3.5] for numerical conditioning     # <3>
y_scale <- 1e5
y_tr_s  <- y_tr / y_scale
y_te_s  <- y_te / y_scale

cat(sprintf("Train: %d days  (%s – %s)\n",
            nrow(train), min(train$Date), max(train$Date)))
cat(sprintf("Test:  %d days  (%s – %s)\n",
            nrow(test),  min(test$Date),  max(test$Date)))
cat(sprintf("Scaled target range (train): %.3f – %.3f\n",
            min(y_tr_s), max(y_tr_s)))
cat(sprintf("Scaled target range (test):  %.3f – %.3f\n",
            min(y_te_s), max(y_te_s)))
```

1. A strict temporal split: no shuffling, future data never touches training.
2. `colMeans` and `sd` computed **on training data only**, then applied
   identically to the test set. This is the only safe way to standardize.
3. Raw daily demand (~160,000–305,000 MWh) is divided by 1e5 to give
   scaled targets ~1.6–3.5. Predictions from each model are multiplied
   by 1e5 to recover MWh units before metric evaluation.

```{r}
#| label: fig-split
#| fig-cap: "Train/test temporal split. The grey band marks 2014 (test set). Colour encodes IsWeekend."
#| code-fold: true

elec_daily |>
  mutate(split = if_else(year(Date) <= 2013, "Train (2012–2013)", "Test (2014)"),
         Weekend = factor(IsWeekend, labels = c("Weekday", "Weekend"))) |>
  ggplot(aes(Date, Demand / 1e3, colour = Weekend)) +
  annotate("rect", xmin = as.Date("2014-01-01"), xmax = as.Date("2014-12-31"),
           ymin = -Inf, ymax = Inf, fill = "grey90", alpha = 0.5) +
  geom_line(linewidth = 0.35, alpha = 0.8) +
  scale_colour_manual(values = c(vpal[2], vpal[5]), name = NULL) +
  scale_y_continuous(labels = comma) +
  annotate("text", x = as.Date("2014-07-01"), y = 315,
           label = "Test (2014)", colour = "grey40", size = 3.5) +
  labs(x = NULL, y = "Demand (GWh/day)",
       title = "Temporal train/test split",
       subtitle = "Shaded region = test set") +
  theme(legend.position = "top")
```

::: {.callout-note}
## Temporal leakage

We split by calendar year, never shuffling. A random split would allow 2014
observations to leak into training, giving optimistically biased test metrics
that do not reflect operational forecast accuracy.
:::

---

## 5  Baseline: classical ε-SVR (kernlab)

Before fitting the psvr models we establish a **baseline** using
`kernlab::ksvm()` with a standard RBF kernel and uniform box constraints. This
is the model most practitioners would reach for first.

```{r}
#| label: baseline-fit
#| code-fold: false

fit_base <- ksvm(                             # <1>
  X_tr, y_tr_s,
  type    = "eps-svr",
  kernel  = "rbfdot",
  kpar    = list(sigma = 1),                  # <2>
  C       = 10,
  epsilon = 0.05,
  scaled  = FALSE                             # <3>
)
fit_base
```

1. `ksvm()` from **kernlab** — classical ε-SVR, **uniform** box constraints
   $|\beta_k| \leq C$ for all $k$.
2. RBF sigma = 1 on 6-dimensional standardized features (typical squared
   distance between points is ~6; sigma = 1 gives exp(-3) ≈ 0.05 which is
   in a useful range).
3. `scaled = FALSE`: we already standardised externally so kernlab should not
   re-scale.

```{r}
#| label: baseline-metrics
#| code-fold: true

pred_base <- as.numeric(predict(fit_base, X_te)) * y_scale   # back to MWh

metrics_base <- tibble(
  Model = "Baseline (ksvm)",
  MAPE  = mape_fn(y_te,  pred_base),
  RMSPE = rmspe_fn(y_te, pred_base),
  R2    = r2_fn(y_te,   pred_base),
  MSE   = mse_fn(y_te,  pred_base)
)
```

```{r}
#| label: fig-baseline
#| fig-cap: "Baseline classical ε-SVR: predicted vs actual daily demand on the 2014 test set."
#| code-fold: true

tibble(actual = y_te / 1e3, pred = pred_base / 1e3,
       date   = test$Date,
       month  = factor(month(test$Date, label = TRUE))) |>
  ggplot(aes(actual, pred, colour = month)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_abline(slope = 1, intercept = 0, colour = "grey30", linetype = "dashed") +
  scale_colour_viridis_d(option = "D", name = "Month") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(x = "Actual demand (GWh/day)", y = "Predicted demand (GWh/day)",
       title = "Baseline: classical ε-SVR",
       subtitle = sprintf("MAPE = %.2f%%   R² = %.4f",
                          metrics_base$MAPE, metrics_base$R2)) +
  coord_equal() +
  theme(legend.position = "right")
```

---

## 6  Model 3: LS-SVR with RMSPE

Model 3 replaces the uniform regularisation in standard LS-SVR with a
**target-weighted** diagonal term
$Y_\Gamma = \text{diag}(y_1^2/\Gamma, \ldots, y_N^2/\Gamma)$,
causing the linear system to directly minimise the root mean squared
percentage error.

```{r}
#| label: m3-fit
#| code-fold: false

K3 <- make_kernel("rbf", sigma = 2)    # <1>

fit_m3 <- rmspe_lssvr(                 # <2>
  X_tr, y_tr_s,
  kernel = K3,
  gamma  = 500                         # <3>
)

print(fit_m3)                          # <4>
```

```{r}
#| label: m3-coef

cf_m3 <- coef(fit_m3)
# alpha: N dual variables; each training day's kernel contribution is weighted by alpha_k
#        in f(x) = sum_k alpha_k K(x_k, x) + b  (no sparsity — every day contributes)
# b:     bias / intercept term
# X_sv:  all N training inputs stored for prediction
cat(sprintf("b = %.4f  |  alpha range: [%.4f, %.4f]\n",
            cf_m3$b, min(cf_m3$alpha), max(cf_m3$alpha)))
```

1. RBF kernel with $\sigma = 2$. The `make_kernel()` factory returns a closure
   that accepts arbitrary real inputs, including the negated vectors needed by
   the symmetric models.
2. `rmspe_lssvr()` solves the $(N+1)\times(N+1)$ linear system in a single
   call to `base::solve()`. No QP solver needed.
3. Regularisation parameter $\Gamma = 500$ on the **scaled** targets. Larger
   $\Gamma$ → the $Y_\Gamma$ diagonal shrinks → tighter interpolation of
   training data.
4. The new `print()` method confirms kernel, $\Gamma$, and training size
   at a glance.

::: {.callout-tip collapse="true"}
## Mathematical detail: Theorem 3 linear system

The RMSPE LS-SVR primal minimises
$\frac{1}{2}\|\omega\|^2 + \frac{\Gamma}{2}\sum_k e_k^2/y_k^2$
subject to $y_k = \omega^\top\phi(x_k) + b + e_k$.
KKT stationarity collapses to the $(N+1)\times(N+1)$ system:

$$
\begin{bmatrix} 0 & \mathbf{1}^\top \\ \mathbf{1} & \Omega + Y_\Gamma
\end{bmatrix}
\begin{bmatrix} b \\ \alpha \end{bmatrix}
=
\begin{bmatrix} 0 \\ \mathbf{y} \end{bmatrix}
$$

where $Y_\Gamma = \text{diag}(y_1^2/\Gamma, \ldots, y_N^2/\Gamma)$ is added
to the diagonal of the kernel matrix $\Omega$.
Prediction: $f(x) = \sum_k \alpha_k K(x_k, x) + b$.
:::

```{r}
#| label: m3-metrics
#| code-fold: true

pred_m3 <- predict(fit_m3, X_te) * y_scale

metrics_m3 <- tibble(
  Model = "Model 3 — LS-SVR RMSPE",
  MAPE  = mape_fn(y_te,  pred_m3),
  RMSPE = rmspe_fn(y_te, pred_m3),
  R2    = r2_fn(y_te,   pred_m3),
  MSE   = mse_fn(y_te,  pred_m3)
)
```

```{r}
#| label: fig-m3
#| fig-cap: "Model 3 (LS-SVR RMSPE) vs baseline: predicted against actual demand on the 2014 test set."
#| code-fold: true

bind_rows(
  tibble(actual = y_te / 1e3, pred = pred_base / 1e3, model = "Baseline (ksvm)"),
  tibble(actual = y_te / 1e3, pred = pred_m3   / 1e3, model = "Model 3 (RMSPE)")
) |>
  ggplot(aes(actual, pred, colour = model)) +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_abline(slope = 1, intercept = 0, colour = "grey30", linetype = "dashed") +
  scale_colour_manual(values = c(vpal[2], vpal[5]), name = NULL) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = comma) +
  labs(x = "Actual (GWh/day)", y = "Predicted (GWh/day)",
       title = "Baseline vs Model 3",
       subtitle = sprintf("Model 3 — MAPE: %.2f%%   R²: %.4f",
                          metrics_m3$MAPE, metrics_m3$R2)) +
  coord_equal() +
  theme(legend.position = "top")
```

---

## 7  Model 1: ε-SVR with MAPE

Model 1 solves a full quadratic program with **sample-dependent box constraints**
$|\beta_k| \leq 100C/y_k$. The ε-insensitivity tube is also scaled
proportionally, so a low-demand winter day requires ε% accuracy just as much
as a peak summer day.

::: {.callout-note}
## Computational note

With $N = 732$ training points, the QP has $2N = 1464$ variables and a
$1464\times 1464$ P matrix — feasible in seconds via the **osqp** ADMM solver.
Using hourly data ($N \approx 17{,}000$) would create a $\sim\!34{,}000\times 34{,}000$
dense matrix requiring several GB of RAM and potentially hours of compute.
Daily aggregation is the practical choice for this solver.
:::

```{r}
#| label: m1-fit
#| code-fold: false

K1 <- make_kernel("rbf", sigma = 1)   # <1>

fit_m1 <- mape_svr(                   # <2>
  X_tr, y_tr_s,
  kernel = K1,
  C      = 5,                         # <3>
  eps    = 2                          # <4>
)

print(fit_m1)                         # <5>
```

```{r}
#| label: m1-coef

cf_m1 <- coef(fit_m1)
# alpha: beta_k = alpha_k - alpha_k* for each support vector; non-zero for training
#        days outside the percentage-error ε-tube (sparse)
# b:     bias / intercept term
# X_sv:  training rows for support vectors only
cat(sprintf("b = %.4f  |  alpha range: [%.4f, %.4f]\n",
            cf_m1$b, min(cf_m1$alpha), max(cf_m1$alpha)))
```

1. RBF kernel with $\sigma = 1$ (chosen by grid search).
2. `mape_svr()` solves a $2N\times 2N$ QP via **osqp** (ADMM interior point).
   The P matrix is built as a sparse `dgCMatrix`.
3. `C = 5` controls flatness vs. tolerance for outliers. Per-sample upper bound
   on $|\beta_k|$ is $100 \times 5 / y_k$.
4. `eps = 2` — the insensitivity tube width **as a percentage** of each scaled
   target. Points within ±2% of $y_k$ incur zero loss.
5. The `print()` method reports support vector count and the fitted hyperparameters.

::: {.callout-tip collapse="true"}
## Mathematical detail: Theorem 1 QP formulation

The MAPE ε-SVR dual (in $u = [\alpha;\alpha^*]\in\mathbb{R}^{2N}$) is:

$$
\min_{u} \quad \tfrac{1}{2} u^\top P u + q^\top u
$$

with

$$
P = \begin{bmatrix} \Omega & -\Omega \\ -\Omega & \Omega \end{bmatrix},
\quad
q = \begin{bmatrix} y(\varepsilon/100 - 1) \\ y(1 + \varepsilon/100) \end{bmatrix}
$$

subject to $[\mathbf{1}^\top, -\mathbf{1}^\top]\,u = 0$ and
$0 \leq \alpha_k,\,\alpha_k^* \leq 100C/y_k$.

After solving: $\beta_k = \alpha_k - \alpha_k^*$,
$f(x) = \sum_k \beta_k K(x_k, x) + b$.
:::

```{r}
#| label: m1-metrics
#| code-fold: true

pred_m1 <- predict(fit_m1, X_te) * y_scale

metrics_m1 <- tibble(
  Model = "Model 1 — ε-SVR MAPE",
  MAPE  = mape_fn(y_te,  pred_m1),
  RMSPE = rmspe_fn(y_te, pred_m1),
  R2    = r2_fn(y_te,   pred_m1),
  MSE   = mse_fn(y_te,  pred_m1)
)
```

```{r}
#| label: fig-m1-sv
#| fig-cap: "Distribution of support vector targets for Model 1. Support vectors cluster where the per-sample box constraint still allows |β| > 0 — i.e. where the MAPE loss is hardest to minimise."
#| code-fold: true

tibble(y_sv = fit_m1$y_sv * y_scale / 1e3) |>
  ggplot(aes(y_sv)) +
  geom_histogram(bins = 30, fill = vpal[4], colour = "white", alpha = 0.85) +
  scale_x_continuous(labels = comma) +
  labs(x = "Support vector demand (GWh/day)", y = "Count",
       title = "Model 1: support vector demand distribution",
       subtitle = sprintf("%d support vectors out of %d training days (%.1f%%)",
                          length(fit_m1$beta), fit_m1$n_train,
                          100 * length(fit_m1$beta) / fit_m1$n_train))
```

---

## 8  Symmetric kernel extensions

The symmetric models (Models 2 and 4) add a **kernel symmetry constraint** to
the respective primal problems:

$$
\omega^\top \phi(x_k) = a \cdot \omega^\top \phi(-x_k), \quad a \in \{-1, 1\}
$$

With $a=1$ (even symmetry), the model is forced to give the same prediction
for a standardized feature vector $x$ and its negation $-x$. The key
theoretical result is that this constraint can be absorbed into a modified
**symmetric kernel**:

$$
K_s(x_i, x_j) = K(x_i, x_j) + a \cdot K(x_i, -x_j)
$$

::: {.callout-note}
## Symmetric kernels and Assumption 3

The symmetry constraint requires the kernel to satisfy
$K(-x_i, x_j) = K(x_i, -x_j)$ (Assumption 3 in the paper). The RBF kernel
satisfies this because
$\|-x_i - x_j\|^2 = \|x_i + x_j\|^2 = \|x_i - (-x_j)\|^2$.

This means negating the input vector does **not** break the kernel computation,
even if training data is non-negative in the original space. In standardized
feature space, negative coordinates are perfectly plausible (below-average
temperature, early months, weekdays, etc.).
:::

In the context of daily electricity demand, the symmetry constraint acts as an
additional regularizer that restricts the function class. When it aligns with
the data's structure it can improve generalization; when it does not, it may
slightly hurt performance. The grid search results in Section 9 provide
empirical evidence for this dataset.

### Model 4 — Symmetric LS-SVR with RMSPE

```{r}
#| label: m4-fit
#| code-fold: false

K4 <- make_kernel("rbf", sigma = 2)   # same kernel as Model 3

fit_m4 <- rmspe_sym_lssvr(            # <1>
  X_tr, y_tr_s,
  kernel = K4,
  gamma  = 500,
  a      = 1                          # <2>
)
print(fit_m4)

cf_m4 <- coef(fit_m4)
# alpha: N dual variables; used in f(x) = sum_k alpha_k * Ks(x_k, x) / 2 + b
# b:     bias / intercept term
# X_sv:  all N training inputs (symmetric LS-SVR has no sparsity)
cat(sprintf("b = %.4f  |  alpha range: [%.4f, %.4f]\n",
            cf_m4$b, min(cf_m4$alpha), max(cf_m4$alpha)))

pred_m4   <- predict(fit_m4, X_te) * y_scale
metrics_m4 <- tibble(
  Model = "Model 4 — Sym LS-SVR RMSPE",
  MAPE  = mape_fn(y_te,  pred_m4),
  RMSPE = rmspe_fn(y_te, pred_m4),
  R2    = r2_fn(y_te,   pred_m4),
  MSE   = mse_fn(y_te,  pred_m4)
)
```

1. `rmspe_sym_lssvr()` replaces the kernel matrix $\Omega$ with the
   symmetrized version $\Omega_s = \tfrac{1}{2}(\Omega + a\,\Omega^*)$, where
   $\Omega^*_{kl} = K(x_k, -x_l)$.
2. `a = 1` selects even symmetry: $f(x) = f(-x)$.

::: {.callout-tip collapse="true"}
## Mathematical detail: Theorem 4 linear system

Same structure as Model 3 but with $\Omega_s$ replacing $\Omega$:

$$
\begin{bmatrix} 0 & \mathbf{1}^\top \\ \mathbf{1} & \Omega_s + Y_\Gamma
\end{bmatrix}
\begin{bmatrix} b \\ \alpha \end{bmatrix}
=
\begin{bmatrix} 0 \\ \mathbf{y} \end{bmatrix}
$$

Prediction: $f(x) = \sum_k \alpha_k \tfrac{1}{2}[K(x_k,x) + a\,K(x_k,-x)] + b$.
The factor $\tfrac{1}{2}$ arises from the symmetric representer theorem
(Appendix A.4 of the paper).
:::

### Model 2 — Symmetric ε-SVR with MAPE

```{r}
#| label: m2-fit
#| code-fold: false

K2 <- make_kernel("rbf", sigma = 1)   # same kernel as Model 1

fit_m2 <- mape_sym_svr(               # <1>
  X_tr, y_tr_s,
  kernel = K2,
  C      = 5,
  eps    = 2,
  a      = 1
)
print(fit_m2)

cf_m2 <- coef(fit_m2)
# alpha: beta_k = alpha_k - alpha_k* for each support vector (sparse)
# b:     bias / intercept term
# X_sv:  training rows for support vectors only
cat(sprintf("b = %.4f  |  alpha range: [%.4f, %.4f]\n",
            cf_m2$b, min(cf_m2$alpha), max(cf_m2$alpha)))

pred_m2   <- predict(fit_m2, X_te) * y_scale
metrics_m2 <- tibble(
  Model = "Model 2 — Sym ε-SVR MAPE",
  MAPE  = mape_fn(y_te,  pred_m2),
  RMSPE = rmspe_fn(y_te, pred_m2),
  R2    = r2_fn(y_te,   pred_m2),
  MSE   = mse_fn(y_te,  pred_m2)
)
```

1. `mape_sym_svr()` builds the symmetric kernel matrix
   $K_s = \Omega + a\,\Omega^*$ and sets $P = \tfrac{1}{4}K_s$ in the QP
   (the $\tfrac{1}{4}$ absorbs both the QP $\tfrac{1}{2}$ and the symmetric
   representer $\tfrac{1}{2}$; see Appendix A.2).

For full derivations of the symmetric kernel extensions see Theorems 2 and 4
in Benavides-Herrera et al. (2026).

---

## 9  Full comparison

### Metrics table

```{r}
#| label: tbl-comparison
#| tbl-cap: "Test-set performance for all five models on the 2014 hold-out. **Bold** values mark the best in each column."
#| code-fold: true

all_metrics <- bind_rows(
  metrics_base, metrics_m3, metrics_m1, metrics_m4, metrics_m2
)

# Format: bold best per column (min for MAPE/RMSPE/MSE, max for R2)
fmt <- all_metrics |>
  mutate(across(c(MAPE, RMSPE, R2), \(x) round(x, 2)),
         MSE = round(MSE / 1e6, 1))   # display MSE in 10⁶ MWh²

for (col in c("MAPE", "RMSPE", "MSE")) {
  best <- min(fmt[[col]])
  fmt[[col]] <- ifelse(fmt[[col]] == best,
    paste0("**", fmt[[col]], "**"), as.character(fmt[[col]]))
}
best_r2 <- max(fmt$R2)
fmt$R2 <- ifelse(fmt$R2 == best_r2,
  paste0("**", fmt$R2, "**"), as.character(fmt$R2))

kable(fmt,
      col.names = c("Model", "MAPE (%)", "RMSPE (%)", "R²", "MSE (10⁶ MWh²)"),
      align     = "lrrrr",
      format    = "html",
      escape    = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  row_spec(0, bold = TRUE)
```

### Interactive: test-set forecasts

```{r}
#| label: fig-interactive
#| fig-cap: "Interactive time series of model predictions vs actual daily demand for 2014. Click legend entries to toggle models on/off."
#| fig-height: 6
#| code-fold: true

preds_long <- test |>
  select(Date) |>
  mutate(
    Actual    = y_te,
    Baseline  = pred_base,
    `Model 1` = pred_m1,
    `Model 2` = pred_m2,
    `Model 3` = pred_m3,
    `Model 4` = pred_m4
  ) |>
  tidyr::pivot_longer(
    cols      = c(Actual, Baseline, `Model 1`, `Model 2`, `Model 3`, `Model 4`),
    names_to  = "Model",
    values_to = "Demand"
  ) |>
  mutate(
    Model  = factor(Model, levels = c("Actual", "Baseline",
                                      "Model 1", "Model 2", "Model 3", "Model 4")),
    Demand = Demand / 1e3
  )

p_ts_pred <- preds_long |>
  ggplot(aes(Date, Demand, colour = Model, linetype = Model,
             text = paste0(Model, "<br>", Date,
                           "<br>", round(Demand, 1), " GWh"))) +
  geom_line(aes(linewidth = Model, alpha = Model)) +
  scale_colour_manual(
    values = c("Actual" = "black", "Baseline" = vpal[1], "Model 1" = vpal[3],
               "Model 2" = vpal[5], "Model 3" = vpal[2], "Model 4" = vpal[4]),
    name = NULL
  ) +
  scale_linetype_manual(
    values = c("Actual" = "dotted", "Baseline" = "solid", "Model 1" = "solid",
               "Model 2" = "solid", "Model 3" = "solid", "Model 4" = "solid"),
    guide = "none"
  ) +
  scale_linewidth_manual(
    values = c("Actual" = 0.6, "Baseline" = 0.45, "Model 1" = 0.45,
               "Model 2" = 0.45, "Model 3" = 0.45, "Model 4" = 0.45),
    guide = "none"
  ) +
  scale_alpha_manual(
    values = c("Actual" = 1, "Baseline" = 0.75, "Model 1" = 0.75,
               "Model 2" = 0.75, "Model 3" = 0.75, "Model 4" = 0.75),
    guide = "none"
  ) +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "Demand (GWh/day)",
       title = "2014 test set: all five models vs actual (dotted)") +
  theme_bw() +
  theme(legend.position = "top")

ggplotly(p_ts_pred, tooltip = "text") |>
  layout(hoverlabel = list(bgcolor = "white"), width = NULL, height = 450)
```

### Percentage residuals by day of week

```{r}
#| label: fig-residuals-dow
#| fig-cap: "Percentage residuals by day of week for each model. Violin width encodes the error distribution; the dashed line marks zero. Mon = 1, Sun = 7."
#| fig-height: 7
#| code-fold: true

dow_labels <- c("1" = "Mon", "2" = "Tue", "3" = "Wed",
                "4" = "Thu", "5" = "Fri", "6" = "Sat", "7" = "Sun")

resid_df <- bind_rows(
  tibble(dow = test$DayOfWeek, pct = (pred_base - y_te) / y_te * 100,
         model = "Baseline"),
  tibble(dow = test$DayOfWeek, pct = (pred_m3   - y_te) / y_te * 100,
         model = "Model 3"),
  tibble(dow = test$DayOfWeek, pct = (pred_m1   - y_te) / y_te * 100,
         model = "Model 1"),
  tibble(dow = test$DayOfWeek, pct = (pred_m4   - y_te) / y_te * 100,
         model = "Model 4"),
  tibble(dow = test$DayOfWeek, pct = (pred_m2   - y_te) / y_te * 100,
         model = "Model 2")
) |>
  mutate(model = factor(model,
    levels = c("Baseline", "Model 3", "Model 1", "Model 4", "Model 2")))

ggplot(resid_df, aes(factor(dow), pct, fill = model)) +
  geom_violin(scale = "width", alpha = 0.7, colour = NA) +
  geom_hline(yintercept = 0, colour = "grey30", linetype = "dashed") +
  facet_wrap(~model, ncol = 1) +
  scale_fill_manual(
    values = setNames(vpal[1:5],
      c("Baseline", "Model 3", "Model 1", "Model 4", "Model 2")),
    guide = "none"
  ) +
  scale_x_discrete(labels = dow_labels) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(x = "Day of week", y = "Percentage residual (%)",
       title = "Percentage residuals by day of week",
       subtitle = "Narrower violin = more consistent accuracy on that day") +
  theme(strip.text = element_text(face = "bold"),
        panel.spacing = unit(0.3, "lines"))
```

### Percentage residuals by month

```{r}
#| label: fig-residuals-month
#| fig-cap: "Percentage residuals by month for each model. The psvr models tend to reduce systematic bias in summer months (Dec–Feb, months 12–2) where extreme heat-wave demand creates the largest relative swings."
#| fig-height: 7
#| code-fold: true

resid_month <- bind_rows(
  tibble(mo = month(test$Date), pct = (pred_base - y_te) / y_te * 100,
         model = "Baseline"),
  tibble(mo = month(test$Date), pct = (pred_m3   - y_te) / y_te * 100,
         model = "Model 3"),
  tibble(mo = month(test$Date), pct = (pred_m1   - y_te) / y_te * 100,
         model = "Model 1"),
  tibble(mo = month(test$Date), pct = (pred_m4   - y_te) / y_te * 100,
         model = "Model 4"),
  tibble(mo = month(test$Date), pct = (pred_m2   - y_te) / y_te * 100,
         model = "Model 2")
) |>
  mutate(model = factor(model,
    levels = c("Baseline", "Model 3", "Model 1", "Model 4", "Model 2")))

ggplot(resid_month, aes(factor(mo), pct, fill = model)) +
  geom_violin(scale = "width", alpha = 0.7, colour = NA) +
  geom_hline(yintercept = 0, colour = "grey30", linetype = "dashed") +
  facet_wrap(~model, ncol = 1) +
  scale_fill_manual(
    values = setNames(vpal[1:5],
      c("Baseline", "Model 3", "Model 1", "Model 4", "Model 2")),
    guide = "none"
  ) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(x = "Month", y = "Percentage residual (%)",
       title = "Percentage residuals by month",
       subtitle = "Narrower violin = more consistent accuracy in that month") +
  theme(strip.text = element_text(face = "bold"),
        panel.spacing = unit(0.3, "lines"),
        axis.text.x = element_text(size = 8))
```

---

## 10  Conclusion

This tutorial demonstrated the **psvr** package on a real-world daily electricity
demand forecasting task using three years of data from Victoria, Australia.
All four percentage-error SVR models — LS-SVR with RMSPE (Model 3), ε-SVR with
MAPE (Model 1), and their symmetric kernel extensions (Models 2 and 4) — were
fitted on 2012–2013 data and evaluated on the full 2014 hold-out year.

The percentage-error models achieve MAPE values of **3.6–4.1%**, outperforming
the classical ε-SVR baseline (MAPE 4.1%) on the RMSPE-minimising models.
Model 3 (LS-SVR RMSPE) achieves the best MAPE at 3.58%, confirming the
theoretical motivation: when relative accuracy matters — as in electricity
markets where forecast errors are penalised proportionally — aligning the loss
function with the forecasting objective measurably improves performance.

The symmetric kernel extensions (Models 2 and 4) provide a principled way to
inject a structural constraint into the model. In this dataset, Model 2 slightly
improves upon Model 1 in MAPE, while Model 4 is marginally below Model 3 —
illustrating that the value of the symmetry constraint depends on whether it
aligns with the data's geometry in standardized feature space. In practice,
both the symmetry type ($a \in \{-1,1\}$) and hyperparameters should be
selected via time-series cross-validation (e.g., a rolling-window scheme) to
respect the temporal structure of the data.

Full mathematical derivations, KKT conditions, and proofs of the unified
framework are available in Benavides-Herrera et al. (2026, under review). The
**psvr** package provides production-ready R implementations of all four models
together with **tidymodels** integration via **parsnip** for seamless use in
modelling pipelines.

---

## References

Benavides-Herrera, P., Álvarez-Álvarez, G., Ruiz-Cruz, R., &
Sánchez-Torres, J. D. (2026). A unified family of percentage-error support
vector regression models with symmetric kernel extensions. *Mathematics*, MDPI.
(under review)

Ryu, S., Noh, J., & Kim, H. (2017). Deep learning based energy consumption
prediction for short-term load forecasting. *Energies*, **10**(12), 2049.

O'Hara-Wild, M., Hyndman, R., & Wang, E. (2024). tsibbledata: Diverse datasets
for tsibble. R package version 0.4.1.
