---
title: "Case Study: Victorian Electricity Demand"
subtitle: "Time series forecasting with psvr, modeltime, and tidymodels"
author: "Pablo Benavides-Herrera"
date: last-modified
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: kable
---

::: {.callout-note}
## Reproducibility note
These results correspond to the manuscript *"A Unified Family of
Percentage-Error Support Vector Regression Models with Symmetric Kernel
Extensions"* submitted to *Mathematics* (MDPI), currently under review.
Model implementations and evaluation protocol match those reported in the
paper. Final published results may differ if revisions are requested.
Tuning results are available as a downloadable file at the end of this
article.
:::

## 1 Introduction

Electricity demand forecasting is a canonical application where relative
accuracy—measured by the Mean Absolute Percentage Error (MAPE)—carries direct
operational significance. Victoria, Australia operates a competitive spot
market in which generators must submit binding supply offers one day ahead; a
5% forecast error on a 200 MWh peak day and a 5% error on an 80 MWh off-peak
day have proportionally equivalent implications for reserve capacity planning
and dispatch scheduling. Absolute-error objectives such as MSE or MAE do not
respect this symmetry, motivating loss functions that explicitly penalise
relative deviations. The four psvr models studied here are derived from
percentage-error loss functions—MAPE and RMSPE—and provide a principled
alternative to standard SVR for this setting.

An additional structural feature of daily electricity demand motivates
exploration of symmetric kernels. After seasonal decomposition and mean
centering, days with similar temperatures on opposite sides of the seasonal
cycle—spring versus autumn—exhibit comparable demand profiles. The symmetric
kernel $K^s(x, x') = K(x, x') + a \cdot K(x, -x')$ captures this reflected
structure explicitly: the cross-term $K(x, -x')$ measures similarity between
$x$ and the reflection of $x'$, so pairs of days that are symmetric under the
seasonal mean are assigned high kernel affinity. Models 2 and 4 in the psvr
family incorporate this symmetrisation.

Temporal causality requires that model selection use rolling-origin
cross-validation (Hyndman & Athanasopoulos, 2021) rather than random splits.
Leaking future observations into any training fold would produce optimistically
biased CV estimates and invalidate the hyperparameter selection step.
A rolling-origin protocol with 10 expanding windows is adopted here, the time
series analog of the multi-seed protocol used in the companion cross-section
case studies. The number of windows is bounded by the available training
horizon: with a 12-month initial training period and 1-month assessment windows
over a 24-month training set, additional folds beyond ten are highly correlated
and contribute diminishing information. In total, 12 models are evaluated: six 
psvr variants (Models 1–4, with LS-RMSPE appearing twice for MAPE- and RMSPE-optimised
 CV selection), three ML baselines (standard RBF-SVR, Random Forest, Linear Regression), 
and three time-series baselines (ARIMAX, ETS, Prophet with regressors).

## 2 Setup

```{r}
#| label: setup

library(tidyverse)
library(tidymodels)
library(modeltime)
library(timetk)
library(psvr)
library(tsibbledata)
library(lubridate)
library(knitr)
library(kableExtra)
library(xfun)
library(patchwork)
library(future)

tidymodels_prefer()
theme_set(theme_bw(base_size = 12))

here::i_am("vignettes/articles/electricity-forecasting.qmd")
```

## 3 Data

The `vic_elec` dataset from **tsibbledata** records half-hourly electricity
demand (MWh) and ambient temperature for Victoria, Australia over 2012–2014.
Observations are aggregated to the daily level by summing demand, averaging
temperature, and flagging any half-hour marked as a holiday. Temperature
squared is appended to allow the recipe to capture the well-documented
U-shaped demand–temperature relationship without requiring manual interaction
terms.

```{r}
#| label: data

elec_daily <- vic_elec |>
  as_tibble() |>
  mutate(Date = as_date(Time)) |>
  group_by(Date) |>
  summarise(
    Demand      = sum(Demand) / 1000,  # MWh -> GWh
    Temperature = mean(Temperature, na.rm = TRUE),
    IsHoliday   = as.integer(any(Holiday)),
    .groups     = "drop"
  ) |>
  mutate(Temperature2 = Temperature^2)

# Scale target to ~[0.5, 1.7] for numerical conditioning.
# The psvr MAPE box constraints are 100*C/y_k — with
# unscaled MWh values these collapse to near zero and
# psvr cannot learn. MAPE/RMSPE are scale-invariant so
# evaluation metrics are unaffected.
y_scale    <- mean(elec_daily$Demand)
elec_daily <- elec_daily |>
  mutate(Demand = Demand / y_scale)

stopifnot(all(elec_daily$Demand > 0))
```

```{r}
#| label: data-summary

tibble(
  Statistic = c(
    "n observations", "First date", "Last date",
    "Min Demand (MWh)", "Median Demand (MWh)",
    "Mean Demand (MWh)", "Max Demand (MWh)"
  ),
  Value = c(
    nrow(elec_daily),
    format(min(elec_daily$Date)),
    format(max(elec_daily$Date)),
    round(min(elec_daily$Demand),    2),
    round(median(elec_daily$Demand), 2),
    round(mean(elec_daily$Demand),   2),
    round(max(elec_daily$Demand),    2)
  )
) |>
  knitr::kable(format = "html", col.names = c("Statistic", "Value"),
               caption = "Victorian daily electricity demand: dataset summary.")
```

```{r}
#| label: ts-plot

plot_time_series(
  elec_daily, Date, Demand,
  .interactive = FALSE,
  .title = "Victorian daily electricity demand (MWh)"
)
```

::: {.callout-tip}
Calendar features (day of week, week of year, month, quarter,
weekend/holiday indicators) are generated automatically from the `Date`
column using `step_timeseries_signature()` in the recipe preprocessing
pipeline. `Temperature^2` is included to capture the nonlinear U-shaped
demand–temperature relationship. ETS is the only purely univariate baseline
and uses a separate recipe without these features.
:::

Daily demand is converted from MWh to GWh and further
divided by the training-set mean, producing values in
approximately [0.5, 1.7]. This scaling is required for
the psvr QP formulation: the MAPE box constraints
$\alpha_k \in [0, 100C/y_k]$ collapse to near zero for
large $y_k$, preventing the solver from learning. MAPE
and RMSPE are scale-invariant so evaluation metrics are
unaffected. All scaled values satisfy $y_k > 0$.

## 4 Train / test split

```{r}
#| label: split

splits <- time_series_split(
  elec_daily,
  date_var   = Date,
  assess     = "1 year",
  cumulative = TRUE
)

train <- training(splits)
test  <- testing(splits)

cat("Training:", nrow(train), "rows |",
    format(min(train$Date)), "to",
    format(max(train$Date)), "\n")
cat("Test:    ", nrow(test),  "rows |",
    format(min(test$Date)),  "to",
    format(max(test$Date)),  "\n")
```

## 5 Rolling-origin cross-validation

Random train/test splits are inappropriate for time series data because they
allow future observations to appear in the training fold, producing
optimistically biased estimates. Rolling-origin CV preserves temporal
causality by always training on the past and validating on the future. With
each successive window the training set expands by one period (cumulative =
`TRUE`), so the model sees progressively more history—mirroring actual
deployment conditions. Ten rolling windows are used, providing stable estimates
of generalisation error while respecting the available
training horizon. With a 24-month training period and
one-month assessment windows, additional folds beyond
ten are highly correlated and contribute diminishing
information.

```{r}
#| label: cv-plan

folds <- time_series_cv(
  train,
  date_var    = Date,
  assess      = "1 month",
  initial     = "12 months",
  slice_limit = 10,
  cumulative  = TRUE
)

cat("CV windows generated:", nrow(folds), "\n")
```

```{r}
#| label: cv-plot
#| fig-width: 10
#| fig-height: 5

folds |>
  tk_time_series_cv_plan() |>
  plot_time_series_cv_plan(
    Date, Demand,
    .title       = "Rolling-origin CV plan (10 windows, 1-month horizon)",
    .interactive = FALSE
  )
```

## 6 Recipe

The preprocessing recipe applies `step_timeseries_signature()` to extract a
comprehensive set of calendar features from the `Date` column—including year,
quarter, month, week-of-year, day-of-week, and weekend indicators—without
requiring manual `mutate()` calls. High-cardinality or redundant variants
(ISO, XTS, POSIX, hour, minute, second, and AM/PM columns) are removed via
`step_rm()` to avoid multicollinearity and reduce dimensionality. Any
remaining nominal predictors (including factor-encoded calendar fields) are
one-hot encoded with `step_dummy()`, and all numeric predictors are
standardised by `step_normalize()` so that SVR kernel distances are not
dominated by scale differences between features. ETS, being a purely
univariate method, uses a separate recipe with only `Date` as the predictor.

```{r}
#| label: recipe

# ── Kernel recipe: psvr + SVR-MSE ────────────────────────────
# Continuous features only — RBF kernel degrades with
# high-dimensional one-hot encoded categorical features.
# DayOfWeek, Month, IsWeekend added manually as numeric.
rec_kernel <- recipe(Demand ~ ., data = train) |>
  update_role(Date, new_role = "ID") |>
  step_mutate(
    DayOfWeek = lubridate::wday(Date, week_start = 1),
    Month     = lubridate::month(Date),
    IsWeekend = as.integer(
      lubridate::wday(Date, week_start = 1) >= 6
    )
  ) |>
  step_normalize(all_numeric_predictors())

# ── Full recipe: RF + LR ─────────────────────────────────────
# Tree-based and linear models benefit from rich calendar
# features generated by step_timeseries_signature().
rec_full <- recipe(Demand ~ ., data = train) |>
  update_role(Date, new_role = "ID") |>
  step_timeseries_signature(Date) |>
  step_rm(matches(
    "(iso$|xts$|posixct$|hour$|minute$|second$|am.pm$)"
  )) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>
  step_normalize(all_numeric_predictors())

# ── Univariate recipe: TS baselines ──────────────────────────
# modeltime engines (auto_arima, ets, prophet) require Date
# as a predictor — update_role("ID") hides it from them.
rec_univariate <- recipe(Demand ~ Date, data = train)

# Compute rbf_sigma from kernel recipe baked training data
train_baked      <- rec_kernel |> prep() |>
  bake(new_data = train)
rbf_sigma_custom <- rbf_sigma_psvr_data(
  train_baked |> select(-Demand, -Date)
)

r <- dials::range_get(rbf_sigma_custom, original = FALSE)
cat("Kernel recipe predictors:", ncol(train_baked) - 2, "\n")
cat("Custom rbf_sigma range (log10): [",
    round(r$lower, 4), ",", round(r$upper, 4), "]\n")
```

## 7 Model specifications

::: {.callout-note}
## Model formulations
For the mathematical derivations of all psvr models, see the companion
manuscript. LS-RMSPE variants appear twice: with MAPE CV (suffix `-M`) and
RMSPE CV (suffix `-R`). The model structure is identical; only the CV
selection metric differs, handled at the `workflow_map` stage.

For the symmetric variants (`m2`, `m4a`), the symmetry type
`sym_type ∈ {"even" (a = 1), "odd" (a = -1)}` is treated as a tunable
hyperparameter and selected jointly with `cost`, `svm_margin`, and
`rbf_sigma` during CV. This brings the electricity grid sizes in line
with the cross-section case studies (m2: 96, m4: 32 combinations).
:::

```{r}
#| label: specs

# ── psvr models ──────────────────────────────────────────────
spec_m1 <- psvr_mape_rbf(
  cost = tune(), svm_margin = tune(), rbf_sigma = tune()
) |> set_engine("psvr")

spec_m2 <- psvr_mape_sym_rbf(
  cost = tune(), svm_margin = tune(), rbf_sigma = tune(),
  sym_type = tune()
) |> set_engine("psvr")

# m3a: LS-RMSPE tuned by MAPE
spec_m3a <- psvr_rmspe_rbf(
  cost = tune(), rbf_sigma = tune()
) |> set_engine("psvr")

# m4a: LS-RMSPE + Sym. Kernel tuned by MAPE
spec_m4a <- psvr_rmspe_sym_rbf(
  cost = tune(), rbf_sigma = tune(),
  sym_type = tune()
) |> set_engine("psvr")

# ── ML baselines ─────────────────────────────────────────────
spec_b1 <- svm_rbf(cost = tune(), rbf_sigma = tune()) |>
  set_engine("kernlab") |>
  set_mode("regression")

spec_b2 <- rand_forest(
  mtry = tune(), trees = 500
) |>
  set_engine("ranger") |>
  set_mode("regression")

spec_b4 <- linear_reg() |> set_engine("lm")

# ── TS baselines ─────────────────────────────────────────────
spec_ts1 <- arima_reg() |> set_engine("auto_arima")

spec_ts2 <- exp_smoothing() |> set_engine("ets")

spec_ts3 <- prophet_reg(
  seasonality_weekly = TRUE,
  seasonality_yearly = TRUE
) |> set_engine("prophet")
```

## 8 Workflow sets

```{r}
#| label: workflow-sets

# psvr + b1_svrmse: kernel recipe
wf_kernel <- workflow_set(
  preproc = list(kernel = rec_kernel),
  models  = list(
    m1        = spec_m1,
    m2        = spec_m2,
    m3a       = spec_m3a,
    m4a       = spec_m4a,
    b1_svrmse = spec_b1
  )
) |>
  psvr_option_add(train_baked |> select(-Demand, -Date))

# RF + LR: full recipe with rich calendar features
wf_full <- workflow_set(
  preproc = list(full = rec_full),
  models  = list(
    b2_rf = spec_b2,
    b4_lm = spec_b4
  )
)

# TS baselines: univariate recipe
wf_ts <- workflow_set(
  preproc = list(univariate = rec_univariate),
  models  = list(
    ts1_arimax  = spec_ts1,
    ts2_ets     = spec_ts2,
    ts3_prophet = spec_ts3
  )
)

wf_all <- bind_rows(wf_kernel, wf_full, wf_ts)

cat("Total workflows:", nrow(wf_all), "\n")
cat("IDs:\n")
print(wf_all$wflow_id)
```

```{r}
#| label: grids

# Explicit hyperparameter grids matching cross-section protocol.
# sigma values correspond to e1071 gamma = c(0.01, 0.1, 1, 10)
# via equivalence gamma = 1/(2*sigma^2)
sigma_vals <- c(0.224, 0.707, 2.236, 7.071)
C_vals     <- c(0.1, 1, 10, 100)
eps_vals   <- c(0.01, 0.1, 1)
sym_vals   <- c("even", "odd")

# m1: cost + svm_margin + rbf_sigma (48 combinations)
grid_mape_svr <- expand.grid(
  cost       = C_vals,
  svm_margin = eps_vals,
  rbf_sigma  = sigma_vals,
  stringsAsFactors = FALSE
)

# m2: m1 grid × sym_type (96 combinations; sym_type tuned jointly)
grid_mape_sym_svr <- expand.grid(
  cost       = C_vals,
  svm_margin = eps_vals,
  rbf_sigma  = sigma_vals,
  sym_type   = sym_vals,
  stringsAsFactors = FALSE
)

# m3a: cost + rbf_sigma (16 combinations)
grid_rmspe <- expand.grid(
  cost      = C_vals,
  rbf_sigma = sigma_vals,
  stringsAsFactors = FALSE
)

# m4a: m3a grid × sym_type (32 combinations; sym_type tuned jointly)
grid_rmspe_sym <- expand.grid(
  cost      = C_vals,
  rbf_sigma = sigma_vals,
  sym_type  = sym_vals,
  stringsAsFactors = FALSE
)

# b1_svrmse: cost + rbf_sigma (16 combinations)
grid_svm <- expand.grid(
  cost      = C_vals,
  rbf_sigma = sigma_vals,
  stringsAsFactors = FALSE
)

# b2_rf: mtry only
grid_rf <- data.frame(mtry = c(2L, 3L, 4L))

cat("Grid sizes:\n")
cat("  SVR-MAPE       (m1):", nrow(grid_mape_svr),     "\n")
cat("  SVR-MAPE+SK    (m2):", nrow(grid_mape_sym_svr), "\n")
cat("  LS-RMSPE       (m3):", nrow(grid_rmspe),        "\n")
cat("  LS-RMSPE+SK    (m4):", nrow(grid_rmspe_sym),    "\n")
cat("  SVR-MSE        (b1):", nrow(grid_svm),          "\n")
cat("  RF             (b2):", nrow(grid_rf),           "\n")
```

## 9 Tuning

::: {.callout-note}
## CV metric for LS-RMSPE variants
`m3b` and `m4b` use RMSE as the CV selection criterion—the closest native
yardstick equivalent to RMSPE. `m3a` and `m4a` use MAPE. The model structure
is identical in both cases; only hyperparameter selection differs.
:::


```{r}
#| label: tuning


results_file <- here::here(
  "vignettes/articles/case-studies/results/electricity-tune-results.rds"
)


if (!file.exists(results_file)) {

  svm_mse_ids  <- "kernel_b1_svrmse"
  rf_ids       <- "full_b2_rf"
  ts_ids       <- c("univariate_ts1_arimax",
                    "univariate_ts2_ets",
                    "univariate_ts3_prophet")
  lm_ids       <- "full_b4_lm"

  ctrl <- control_grid(
    save_pred     = TRUE,
    parallel_over = "everything",
    verbose       = FALSE
  )

  t0 <- proc.time()
  plan(multisession)
  set.seed(2024)

  tune_all <- tryCatch({
    tune_svr_mape <- wf_all |>
      filter(wflow_id == "kernel_m1") |>
      workflow_map(
        fn        = "tune_grid",
        resamples = folds,
        grid      = grid_mape_svr,
        metrics   = metric_set(mape, rmse, rsq),
        control   = ctrl
      )

    tune_svr_mape_sym <- wf_all |>
      filter(wflow_id == "kernel_m2") |>
      workflow_map(
        fn        = "tune_grid",
        resamples = folds,
        grid      = grid_mape_sym_svr,
        metrics   = metric_set(mape, rmse, rsq),
        control   = ctrl
      )

    tune_ls_rmspe <- wf_all |>
      filter(wflow_id == "kernel_m3a") |>
      workflow_map(
        fn        = "tune_grid",
        resamples = folds,
        grid      = grid_rmspe,
        metrics   = metric_set(mape, rmse, rsq),
        control   = ctrl
      )

    tune_ls_rmspe_sym <- wf_all |>
      filter(wflow_id == "kernel_m4a") |>
      workflow_map(
        fn        = "tune_grid",
        resamples = folds,
        grid      = grid_rmspe_sym,
        metrics   = metric_set(mape, rmse, rsq),
        control   = ctrl
      )

    tune_svm_mse <- wf_all |>
      filter(wflow_id %in% svm_mse_ids) |>
      workflow_map(
        fn        = "tune_grid",
        resamples = folds,
        grid      = grid_svm,
        metrics   = metric_set(mape, rmse, rsq),
        control   = ctrl
      )

    tune_rf <- wf_all |>
      filter(wflow_id %in% rf_ids) |>
      workflow_map(
        fn        = "tune_grid",
        resamples = folds,
        grid      = grid_rf,
        metrics   = metric_set(mape, rmse, rsq),
        control   = ctrl
      )

    tune_ts <- wf_all |>
      filter(wflow_id %in% ts_ids) |>
      workflow_map(
        fn        = "fit_resamples",
        resamples = folds,
        metrics   = metric_set(mape, rmse, rsq),
        control   = control_resamples(save_pred = TRUE)
      )

    tune_lm <- wf_all |>
      filter(wflow_id %in% lm_ids) |>
      workflow_map(
        fn        = "fit_resamples",
        resamples = folds,
        metrics   = metric_set(mape, rmse, rsq),
        control   = control_resamples(save_pred = TRUE)
      )

    elapsed <- round((proc.time() - t0)[["elapsed"]] / 60, 1)
    message(sprintf("[tuning] completed in %.1f min", elapsed))
    bind_rows(tune_svr_mape, tune_svr_mape_sym,
              tune_ls_rmspe, tune_ls_rmspe_sym,
              tune_svm_mse, tune_rf,
              tune_ts, tune_lm)

  }, error = function(e) {
    message("Tuning failed: ", conditionMessage(e))
    NULL
  }, finally = {
    plan(sequential)
  })

  if (!is.null(tune_all)) {
    saveRDS(tune_all, results_file)
  }

} else {
  tune_all <- readRDS(results_file)
}

cat("Tuning results loaded for",
    nrow(tune_all), "workflows.\n")

  # Show which workflows have errors
  tune_status <- tune_all |>
    mutate(
      n_results = purrr::map_int(result, function(r) {
        tryCatch(nrow(collect_metrics(r)), error = function(e) 0L)
      }),
      has_error = n_results == 0L
    ) |>
    select(wflow_id, n_results, has_error)

  print(tune_status)
  if (any(tune_status$has_error)) {
    warning("Some workflows have no results: ",
            paste(tune_status$wflow_id[tune_status$has_error],
                  collapse = ", "))
  }
```

## 9.1 Export per-fold results

```{r}
#| label: export-results

# ── Step 1: Model registry ────────────────────────────────────────────
model_registry <- tribble(
  ~wflow_id,                ~tune_id,                 ~select_metric,  ~model_id, ~label,                                  ~abbrev,         ~family,
  "kernel_m1",              "kernel_m1",              "mape",          "m1",      "SVR-MAPE",                              "SVR-MAPE",      "psvr",
  "kernel_m2",              "kernel_m2",              "mape",          "m2",      "SVR-MAPE + Sym. Kernel",                "SVR-MAPE+SK",   "psvr",
  "kernel_m3a",             "kernel_m3a",             "mape",          "m3a",     "LS-RMSPE (MAPE opt.)",                  "LS-RMSPE-M",    "psvr",
  "kernel_m3a",             "kernel_m3a",             "rmse",          "m3b",     "LS-RMSPE (RMSPE opt.)",                 "LS-RMSPE-R",    "psvr",
  "kernel_m4a",             "kernel_m4a",             "mape",          "m4a",     "LS-RMSPE + Sym. Kernel (MAPE opt.)",    "LS-RMSPE+SK-M", "psvr",
  "kernel_m4a",             "kernel_m4a",             "rmse",          "m4b",     "LS-RMSPE + Sym. Kernel (RMSPE opt.)",   "LS-RMSPE+SK-R", "psvr",
  "kernel_b1_svrmse",       "kernel_b1_svrmse",       "mape",          "b1",      "ε-SVR (MSE)",                           "SVR-MSE",       "baseline",
  "full_b2_rf",             "full_b2_rf",             "mape",          "b2",      "Random Forest",                         "RF",            "baseline",
  "full_b4_lm",             "full_b4_lm",             NA_character_,   "b4",      "Linear Regression",                     "LR",            "baseline",
  "univariate_ts1_arimax",  "univariate_ts1_arimax",  NA_character_,   "ts1",     "ARIMAX",                                "ARIMAX",        "baseline",
  "univariate_ts2_ets",     "univariate_ts2_ets",     NA_character_,   "ts2",     "ETS",                                   "ETS",           "baseline",
  "univariate_ts3_prophet", "univariate_ts3_prophet", NA_character_,   "ts3",     "Prophet + regressors",                  "Prophet",       "baseline"
)

# ── Step 2: Metric helpers (exact formulas from experiment_helpers.R) ─
mape_fn  <- function(y, yhat) mean(abs((y - yhat) / y)) * 100
rmspe_fn <- function(y, yhat) sqrt(mean(((y - yhat) / y)^2)) * 100
maape_fn <- function(y, yhat) mean(atan(abs((y - yhat) / y))) * (200 / pi)
mse_fn   <- function(y, yhat) mean((y - yhat)^2)
r2_fn    <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

mase_lag7 <- function(actual, predicted, train_actual) {
  mae   <- mean(abs(actual - predicted))
  denom <- mean(abs(diff(train_actual, lag = 7L)))
  mae / denom
}

# ── Step 3: Fold actuals lookup ────────────────────────────────────────
# Diagnostic confirmed: collect_predictions() uses zero-padded ids
# (Slice01..Slice10) and absolute .row indices in train (not 1-indexed
# within each fold). Both must match here for the join to work.
fold_actuals <- purrr::imap(folds$splits, function(spl, i) {
  asmnt         <- assessment(spl)
  asmnt_indices <- rsample::complement(spl)
  stopifnot(length(asmnt_indices) == nrow(asmnt))
  tibble(
    id           = sprintf("Slice%02d", i),
    fold_index   = i,
    actual       = asmnt$Demand,
    .row         = asmnt_indices,
    train_demand = list(analysis(spl)$Demand)
  )
}) |> bind_rows()

# Verify join keys match before proceeding
res_check   <- extract_workflow_set_result(tune_all, "kernel_m1")
preds_check <- collect_predictions(res_check)
id_mismatch <- !any(unique(preds_check$id) %in% unique(fold_actuals$id))
if (id_mismatch) stop(
  "ID format mismatch between predictions and fold_actuals. ",
  "Prediction ids: ", paste(unique(preds_check$id)[1:3], collapse = ", "),
  " — Fold actuals ids: ", paste(unique(fold_actuals$id)[1:3], collapse = ", ")
)

# ── Step 4: Fold-prediction extractor ─────────────────────────────────
get_fold_preds <- function(wflow_id, tune_id, select_metric, ...) {
  res <- extract_workflow_set_result(tune_all, tune_id)
  if (is.na(select_metric)) {
    collect_predictions(res)
  } else {
    best <- select_best(res, metric = select_metric)
    collect_predictions(res) |>
      filter(.config == best$.config)
  }
}

# ── Steps 5–6: Per-model per-fold metrics, assemble, save ─────────────
results_file <- here::here(
  "vignettes/articles/case-studies/results/electricity-results.csv"
)


if (!file.exists(results_file)) {

  n_folds   <- nrow(folds)
  slice_ids <- sprintf("Slice%02d", seq_len(n_folds))

  results <- purrr::pmap_dfr(model_registry,
    function(wflow_id, tune_id, select_metric,
             model_id, label, abbrev, family) {

      fold_preds <- tryCatch(
        get_fold_preds(wflow_id, tune_id, select_metric),
        error = function(e) {
          message(sprintf("[export-results] %s failed: %s",
                          model_id, conditionMessage(e)))
          NULL
        }
      )

      if (is.null(fold_preds)) {
        return(tibble(
          dataset  = "electricity",
          seed     = seq_len(n_folds),
          model_id = model_id, label = label,
          abbrev   = abbrev,   family = family,
          MAPE = NA_real_, RMSPE = NA_real_,
          MAAPE = NA_real_, MASE = NA_real_,
          MSE  = NA_real_, R2   = NA_real_
        ))
      }

      fold_data <- fold_preds |>
        inner_join(fold_actuals, by = c("id", ".row"))

      if (nrow(fold_data) == 0L) {
        warning(sprintf(
          "[export-results] zero rows after join for %s", model_id))
        return(tibble(
          dataset  = "electricity",
          seed     = seq_len(n_folds),
          model_id = model_id, label = label,
          abbrev   = abbrev,   family = family,
          MAPE = NA_real_, RMSPE = NA_real_,
          MAAPE = NA_real_, MASE = NA_real_,
          MSE  = NA_real_, R2   = NA_real_
        ))
      }

      fold_data |>
        group_by(id, fold_index) |>
        summarise(
          MAPE  = mape_fn(actual,  .pred),
          RMSPE = rmspe_fn(actual, .pred),
          MAAPE = maape_fn(actual, .pred),
          MASE  = mase_lag7(actual, .pred, train_demand[[1]]),
          MSE   = mse_fn(actual,   .pred),
          R2    = r2_fn(actual,    .pred),
          .groups = "drop"
        ) |>
        mutate(
          dataset  = "electricity",
          seed     = fold_index,
          model_id = model_id, label = label,
          abbrev   = abbrev,   family = family
        ) |>
        select(dataset, seed, model_id, label, abbrev, family,
               MAPE, RMSPE, MAAPE, MASE, MSE, R2)
    }
  )

  write_csv(results, results_file)

} else {
  results <- read_csv(results_file, show_col_types = FALSE)
}

# Verification
cat("Rows:", nrow(results), "| Expected: 120\n")
cat("NA count:", sum(is.na(results$MAPE)), "\n")
print(results |>
  group_by(model_id, abbrev) |>
  summarise(mean_MAPE = round(mean(MAPE, na.rm = TRUE), 3),
            .groups = "drop") |>
  arrange(mean_MAPE))
```

## 9.2 Selected hyperparameters

A summary of hyperparameters selected by CV is reported below. The non-symmetric and symmetric LS-RMSPE variants converge to identical $(C, \sigma)$, isolating the contribution of the symmetric kernel from the tuning process.

```{r}
#| label: hyperparameters-table

hp_registry <- tibble::tribble(
  ~wflow_id,           ~select_metric, ~label,                              ~has_tuning,
  "kernel_m1",         "mape",         "SVR-MAPE",                          TRUE,
  "kernel_m2",         "mape",         "SVR-MAPE + Sym. Kernel",            TRUE,
  "kernel_m3a",        "mape",         "LS-RMSPE",                          TRUE,
  "kernel_m4a",        "mape",         "LS-RMSPE + Sym. Kernel",            TRUE,
  "kernel_b1_svrmse",  "mape",         "ε-SVR (MSE)",                       TRUE,
  "full_b2_rf",        "mape",         "Random Forest",                     TRUE
)

best_hp <- purrr::pmap_dfr(
  hp_registry |> dplyr::filter(has_tuning),
  function(wflow_id, select_metric, label, has_tuning) {
    res  <- extract_workflow_set_result(tune_all, wflow_id)
    best <- select_best(res, metric = select_metric)
    cv   <- collect_metrics(res) |>
      dplyr::filter(.config == best$.config, .metric == "mape") |>
      dplyr::slice(1)
    tibble::tibble(
      Model       = label,
      C           = if ("cost"       %in% names(best)) best$cost       else NA_real_,
      epsilon     = if ("svm_margin" %in% names(best)) best$svm_margin else NA_real_,
      sigma       = if ("rbf_sigma"  %in% names(best)) best$rbf_sigma  else NA_real_,
      gamma       = if ("rbf_sigma"  %in% names(best)) 1 / (2 * best$rbf_sigma^2) else NA_real_,
      sym_type    = if ("sym_type"   %in% names(best)) as.character(best$sym_type) else NA_character_,
      mtry        = if ("mtry"       %in% names(best)) best$mtry       else NA_integer_,
      cv_MAPE_pct = cv$mean,
      cv_SE       = cv$std_err
    )
  }
)

best_hp |>
  dplyr::mutate(
    `C`           = ifelse(is.na(C),        "—", format(C)),
    `ε`           = ifelse(is.na(epsilon),  "—", format(epsilon)),
    `σ`           = ifelse(is.na(sigma),    "—", sprintf("%.3f", sigma)),
    `γ`           = ifelse(is.na(gamma),    "—", sprintf("%.3f", gamma)),
    `Symmetry`    = ifelse(is.na(sym_type), "—", sym_type),
    `mtry`        = ifelse(is.na(mtry),     "—", as.character(mtry)),
    `CV MAPE (%)` = sprintf("%.3f ± %.3f", cv_MAPE_pct, cv_SE)
  ) |>
  dplyr::select(Model, C, `ε`, `σ`, `γ`, Symmetry, mtry, `CV MAPE (%)`) |>
  knitr::kable(
    format  = "html",
    align   = c("l", rep("r", 7)),
    caption = paste(
      "Hyperparameters selected by 10-fold rolling-origin CV.",
      "γ = 1/(2σ²) gives the equivalent RBF gamma in the e1071/sklearn parameterisation.",
      "LS-RMSPE and LS-RMSPE + Sym. Kernel select identical (C, σ),",
      "isolating the contribution of the symmetric kernel structure."
    )
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) |>
  kableExtra::footnote(
    general = paste(
      "Linear Regression and TS baselines have no tunable hyperparameters",
      "in this experiment and are omitted from this table."
    ),
    general_title = "Note:"
  )

hp_results_file <- here::here(
  "vignettes/articles/case-studies/results/electricity-best-hp.csv"
)
if (!file.exists(hp_results_file)) {
  readr::write_csv(best_hp, hp_results_file)
}
```

## 10 Results

### 10.1 CV ranking plot

```{r}
#| label: ranking-plot
#| fig-width: 11
#| fig-height: 6

  # Remove workflows with no valid results before ranking
  tune_valid <- tune_all |>
    filter(purrr::map_lgl(result, function(r) {
      tryCatch({
        nrow(collect_metrics(r)) > 0
      }, error = function(e) FALSE)
    }))

  rank_results(tune_valid, rank_metric = "mape",
               select_best = TRUE) |>
  filter(.metric == "mape") |>
  mutate(
    family = case_when(
      str_detect(wflow_id, "^kernel_m")             ~ "psvr",
      str_detect(wflow_id, "^(kernel_b|full_b)")    ~ "ML baseline",
      str_detect(wflow_id, "^univariate_ts")        ~ "TS baseline"
    )
  ) |>
  ggplot(aes(x = rank, y = mean,
             colour = family, label = wflow_id)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = mean - 1.96 * std_err,
        ymax = mean + 1.96 * std_err),
    width = 0.3
  ) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20) +
  scale_colour_manual(
    values = c("psvr"        = "#2c7bb6",
               "ML baseline" = "#d7191c",
               "TS baseline" = "#1a9641")
  ) +
  labs(x      = "Rank",
       y      = "CV MAPE (%)",
       title  = "Model ranking by CV MAPE (10 rolling windows)",
       colour = "Family") +
  theme(legend.position = "bottom")
```

### 10.2 CV summary table

```{r}
#| label: cv-table

options(knitr.kable.NA = "—")

model_labels <- tibble::tribble(
  ~wflow_id,                    ~label,
  "kernel_m1",                  "SVR-MAPE",
  "kernel_m2",                  "SVR-MAPE + Sym. Kernel",
  "kernel_m3a",                 "LS-RMSPE (MAPE opt.)",
  "kernel_m3b",                 "LS-RMSPE (RMSPE opt.)",
  "kernel_m4a",                 "LS-RMSPE + Sym. Kernel (MAPE opt.)",
  "kernel_m4b",                 "LS-RMSPE + Sym. Kernel (RMSPE opt.)",
  "kernel_b1_svrmse",           "ε-SVR (MSE)",
  "full_b2_rf",                 "Random Forest",
  "full_b4_lm",                 "Linear Regression",
  "univariate_ts1_arimax",      "ARIMAX",
  "univariate_ts2_ets",         "ETS",
  "univariate_ts3_prophet",     "Prophet + regressors"
)

family_labels <- tibble::tribble(
  ~wflow_id,                    ~family,
  "kernel_m1",                  "psvr",
  "kernel_m2",                  "psvr",
  "kernel_m3a",                 "psvr",
  "kernel_m3b",                 "psvr",
  "kernel_m4a",                 "psvr",
  "kernel_m4b",                 "psvr",
  "kernel_b1_svrmse",           "ML baseline",
  "full_b2_rf",                 "ML baseline",
  "full_b4_lm",                 "ML baseline",
  "univariate_ts1_arimax",      "TS baseline",
  "univariate_ts2_ets",         "TS baseline",
  "univariate_ts3_prophet",     "TS baseline"
)

# tune_all only has 10 workflows (m3b/m4b share with m3a/m4a)
# Use tune_valid for CV table (excludes failed workflows)
cv_labels <- model_labels |>
  filter(!wflow_id %in% c("kernel_m3b", "kernel_m4b"))
cv_families <- family_labels |>
  filter(!wflow_id %in% c("kernel_m3b", "kernel_m4b"))

cv_summary <- collect_metrics(tune_valid) |>
  filter(.metric == "mape") |>
  group_by(wflow_id) |>
  slice_min(mean, n = 1) |>
  ungroup() |>
  left_join(cv_labels,    by = "wflow_id") |>
  left_join(cv_families,  by = "wflow_id") |>
  arrange(mean)

cv_summary |>
  mutate(
    `CV MAPE (%)` = sprintf("%.3f ± %.3f",
                            mean, std_err)
  ) |>
  select(Model = label, Family = family,
         `CV MAPE (%)`) |>
  knitr::kable(
    format  = "html",
    escape  = FALSE,
    caption = paste(
      "Mean CV MAPE over 10 rolling-origin windows.",
      "Values shown as mean ± standard error.",
      "psvr models highlighted in blue."
    )
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) |>
  kableExtra::row_spec(
    which(cv_summary$family == "psvr"),
    background = "#eaf4fb"
  )
```

### 10.3 Final fit and test-set forecast

```{r}
#| label: final-fits

# Helper: finalize workflow and refit on full train set
finalize_and_fit <- function(wflow_id_str,
                              metric = "mape") {
  wf  <- extract_workflow(wf_all, id = wflow_id_str)
  res <- extract_workflow_set_result(tune_all,
                                      wflow_id_str)
  # fit_resamples results have no tunable params —
  # use workflow directly without select_best
  has_params <- tryCatch({
    best <- select_best(res, metric = metric)
    TRUE
  }, error = function(e) FALSE)

  if (has_params) {
    best <- select_best(res, metric = metric)
    wf   <- finalize_workflow(wf, best)
  }
  tryCatch(
    last_fit(wf, splits,
             metrics = metric_set(mape, rmse, rsq)),
    error = function(e) {
      warning(sprintf(
        "last_fit failed for %s: %s",
        wflow_id_str, conditionMessage(e)))
      NULL
    }
  )
}

# No-tune models: fit directly
fit_b4 <- extract_workflow(wf_all, "full_b4_lm") |>
  last_fit(splits, metrics = metric_set(mape, rmse, rsq))

fit_ets <- extract_workflow(wf_all,
                            "univariate_ts2_ets") |>
  last_fit(splits, metrics = metric_set(mape, rmse, rsq))

fits <- list(
  kernel_m1              = finalize_and_fit("kernel_m1",
                                             "mape"),
  kernel_m2              = finalize_and_fit("kernel_m2",
                                             "mape"),
  kernel_m3a             = finalize_and_fit("kernel_m3a",
                                             "mape"),
  kernel_m3b             = finalize_and_fit("kernel_m3a",
                                             "rmse"),
  kernel_m4a             = finalize_and_fit("kernel_m4a",
                                             "mape"),
  kernel_m4b             = finalize_and_fit("kernel_m4a",
                                             "rmse"),
  kernel_b1_svrmse       = finalize_and_fit(
                             "kernel_b1_svrmse", "mape"),
  full_b2_rf             = finalize_and_fit("full_b2_rf",
                                             "mape"),
  full_b4_lm             = fit_b4,
  univariate_ts1_arimax  = finalize_and_fit(
                             "univariate_ts1_arimax",
                             "mape"),
  univariate_ts2_ets     = fit_ets,
  univariate_ts3_prophet = finalize_and_fit(
                             "univariate_ts3_prophet",
                             "mape")
)
```

### 10.4 Test-set accuracy table

```{r}
#| label: test-accuracy

test_metrics <- purrr::imap(fits, function(fit, id) {
  if (is.null(fit)) return(NULL)
  metrics <- tryCatch(collect_metrics(fit),
                      error = function(e) NULL)
  if (is.null(metrics) || nrow(metrics) == 0L)
    return(NULL)
  metrics |> mutate(wflow_id = id)
}) |>
  purrr::compact() |>
  dplyr::bind_rows() |>
  left_join(model_labels,  by = "wflow_id") |>
  left_join(family_labels, by = "wflow_id") |>
  select(wflow_id, label, family, .metric, .estimate) |>
  pivot_wider(names_from  = .metric,
              values_from = .estimate) |>
  arrange(mape)

test_metrics |>
  select(Model = label, Family = family,
         `MAPE (%)` = mape,
         `RMSE`     = rmse,
         `R²`       = rsq) |>
  mutate(across(where(is.numeric), ~round(.x, 4))) |>
  knitr::kable(
    format  = "html",
    escape  = FALSE,
    caption = paste(
      "Test-set accuracy on held-out 2014 data, sorted by MAPE.",
      "LS-RMSPE (MAPE opt.) and LS-RMSPE (RMSPE opt.) share the",
      "same model structure — identical values reflect that both",
      "CV criteria selected the same hyperparameters on this dataset.",
      "Same applies to LS-RMSPE + Sym. Kernel variants."
    )
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) |>
  kableExtra::row_spec(
    which(test_metrics$family == "psvr"),
    background = "#eaf4fb"
  )
```

### 10.5 MASE with lag-7 denominator

```{r}
#| label: mase-lag7

# For electricity demand, MASE uses lag-7 (weekly seasonal
# naive) as denominator — standard for daily data with
# weekly seasonality (Hyndman & Koehler, 2006)
mase_lag7 <- function(actual, predicted, train_actual) {
  mae   <- mean(abs(actual - predicted))
  denom <- mean(abs(diff(train_actual, lag = 7L)))
  mae / denom
}

y_train <- train$Demand
y_test  <- test$Demand

mase_results <- purrr::imap(fits, function(fit, id) {
  if (is.null(fit)) return(NULL)
  preds <- tryCatch(
    collect_predictions(fit)$.pred,
    error = function(e) NULL
  )
  if (is.null(preds)) return(NULL)
  tibble(
    wflow_id  = id,
    MASE_lag7 = mase_lag7(y_test, preds, y_train)
  )
}) |>
  purrr::compact() |>
  dplyr::bind_rows() |>
  left_join(model_labels, by = "wflow_id") |>
  arrange(MASE_lag7)

mase_results |>
  select(Model = label, `MASE (lag-7)` = MASE_lag7) |>
  mutate(`MASE (lag-7)` = round(`MASE (lag-7)`, 4)) |>
  knitr::kable(
    format  = "html",
    caption = paste(
      "MASE with weekly seasonal naive denominator (lag=7).",
      "Values < 1 indicate the model outperforms",
      "the seasonal naive baseline."
    )
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

### 10.6 Forecast plot

```{r}
#| label: forecast-plot
#| fig-width: 12
#| fig-height: 7

# Combine all test predictions for plotting
all_preds <- purrr::imap(fits, function(fit, id) {
  if (is.null(fit)) return(NULL)
  tryCatch(
    collect_predictions(fit) |>
      mutate(wflow_id = id,
             Date     = test$Date[.row]),
    error = function(e) NULL
  )
}) |>
  purrr::compact() |>
  dplyr::bind_rows() |>
  left_join(model_labels, by = "wflow_id")

# Plot actual vs predicted for all models
ggplot() +
  geom_line(
    data = test,
    aes(x = Date, y = Demand),
    colour = "black", linewidth = 0.8, alpha = 0.7
  ) +
  geom_line(
    data = all_preds,
    aes(x = Date, y = .pred,
        colour = label),
    linewidth = 0.5, alpha = 0.7
  ) +
  scale_colour_viridis_d(option = "turbo") +
  labs(
    x      = NULL,
    y      = "Daily Demand (MWh)",
    title  = "Test-set forecasts: all 12 models vs. actual (2014)",
    colour = "Model"
  ) +
  theme(legend.position = "bottom",
        legend.text     = element_text(size = 8)) +
  guides(colour = guide_legend(nrow = 3))
```

## 11 Discussion

The results on Victorian electricity demand reveal several patterns consistent
with the theoretical motivation for the psvr models. Symmetric kernel variants
(Models 2 and 4) consistently outperform their non-symmetric counterparts
under MAPE when the seasonal symmetry structure is present in the data:
$K^s(x, x') = K(x, x') + K(x, -x')$ assigns higher affinity to feature
vectors that are reflections of one another under the seasonal mean, directly
exploiting the spring–autumn demand symmetry described in the introduction.

The comparison with time-series baselines requires careful interpretation
because the models do not operate on equal information. ARIMAX and Prophet
both receive temperature and calendar covariates alongside the time index,
while ETS is strictly univariate. This information asymmetry means that ETS
performance represents a natural floor—the baseline achievable from the
temporal pattern alone—whereas ARIMAX and Prophet already benefit from the
same feature-engineering advantages as psvr. That psvr symmetric variants
remain competitive against ARIMAX and Prophet despite having no explicit
temporal structure (no AR components, no trend or seasonality decomposition)
underscores the practical value of MAPE-optimal loss alignment combined with
expressive kernel functions.

MAPE and MASE with a lag-7 seasonal naive denominator generally tell a
consistent story across models, with rankings that agree within their
respective standard errors. Discrepancies arise primarily for models whose
loss function is aligned with neither metric—the MSE-SVR baseline, for
example, often ranks differently under MAPE than under MASE because its
training objective does not penalise relative errors.

The principal trade-off of psvr relative to the automatic TS methods is
engineering overhead: the psvr pipeline requires feature extraction,
preprocessing, kernel selection, and CV-based hyperparameter search, whereas
ETS, ARIMA, and Prophet are largely self-configuring. When temperature and
calendar features are reliably available and the operational metric is MAPE or
a relative-error norm, SVR-MAPE with symmetric kernels is a principled,
competitive alternative. In contexts where feature availability is uncertain
or rapid prototyping is prioritised, automatic TS methods retain an advantage.

## 12 Downloadable files

```{r}
#| label: downloads

xfun::embed_file(
  here::here("vignettes/articles/case-studies/results/electricity-tune-results.rds"),
  text = "Download tuning results (.rds)"
)
```

## References

Benavides-Herrera, P., Álvarez-Álvarez, G., Ruiz-Cruz, R., &
Sánchez-Torres, J. D. (2026). A unified family of percentage-error support
vector regression models with symmetric kernel extensions. *Mathematics*,
MDPI. <https://doi.org/10.5281/zenodo.19643526>

Hyndman, R. J., & Athanasopoulos, G. (2021). *Forecasting: Principles and
Practice* (3rd ed.). OTexts. <https://otexts.com/fpp3/>

Hyndman, R. J., & Koehler, A. B. (2006). Another look at measures of
forecast accuracy. *International Journal of Forecasting*, 22(4), 679–688.

O'Hara-Wild, M., Hyndman, R., & Wang, E. (2024). tsibbledata: Diverse
datasets for tsibble. R package version 0.4.1.
