---
title: "Case Study: Building Energy Efficiency"
subtitle: "Benchmarking psvr models with percentage-error losses"
author: "Pablo Benavides-Herrera"
date: last-modified
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    df-print: kable
---

::: {.callout-note}
## Reproducibility note
These results correspond to the manuscript *"A Unified Family of
Percentage-Error Support Vector Regression Models with Symmetric
Kernel Extensions"* submitted to *Mathematics* (MDPI), currently
under review. Hyperparameter grids, random seeds, and model
implementations match those reported in the paper exactly.
Final published results may differ if revisions are requested.
Source code and raw results are available as downloadable files
at the end of this article.
:::

## 1 Introduction

The Energy Efficiency dataset (Tsanas & Xifara, 2012; UCI ML Repository)
contains n = 768 building simulations with 8 architectural features
(relative compactness, surface area, wall area, roof area, overall
height, orientation, glazing area, glazing area distribution). The
target is heating load Y1 (kWh/m²), strictly positive, ranging from
6.01 to 43.10.

In building energy certification (e.g., LEED, BREEAM, NOM-020),
compliance thresholds are expressed as percentage deviations from
design values, making MAPE the natural evaluation metric: a 10% error
on a low-load building and a 10% error on a high-load building have
proportionally equivalent design implications.

Features X6 (Orientation) and X8 (Glazing Area Distribution) are
ordinal and treated as continuous following standard practice for
kernel and tree-based methods. This article demonstrates that `psvr`
models — trained directly under MAPE and RMSPE loss — provide
competitive percentage-error performance over 30 randomized train/test
splits. Data loaded from the UCI ML Repository.

## 2 Setup

```{r}
#| label: setup

library(tidyverse)
library(tidymodels)
library(psvr)
library(readxl)
library(kernlab)
library(ranger)
library(xgboost)
library(quantreg)
library(e1071)
library(knitr)
library(kableExtra)
library(xfun)
library(patchwork)

source("case-studies/experiment_helpers.R")

tidymodels_prefer()
theme_set(theme_bw(base_size = 12))
```

## 3 Data

```{r}
#| label: data

ee_local <- "case-studies/results/ENB2012_data.xlsx"
if (!file.exists(ee_local)) {
  download.file(
    paste0("https://archive.ics.uci.edu/ml/",
           "machine-learning-databases/00242/ENB2012_data.xlsx"),
    ee_local, mode = "wb", quiet = TRUE)
}
df_ee <- readxl::read_excel(ee_local)
y <- df_ee$Y1
X <- df_ee |> select(X1:X8) |> as.matrix()

stopifnot(all(y > 0))
```

```{r}
#| label: data-summary
#| echo: false

tibble::tibble(
  Statistic = c("n", "p", "min(y)", "median(y)", "mean(y)", "max(y)"),
  Value     = c(
    nrow(X), ncol(X),
    min(y), median(y), mean(y), max(y)
  )
) |>
  mutate(Value = round(Value, 2)) |>
  knitr::kable(caption = "Energy Efficiency dataset summary (heating load Y1).")
```

```{r}
#| label: data-histogram
#| echo: false
#| fig-width: 7
#| fig-height: 4

ggplot(data.frame(y = y), aes(x = y)) +
  geom_histogram(binwidth = 2, fill = "#2c7bb6", color = "white",
                 alpha = 0.85) +
  labs(x = "Heating Load Y1 (kWh/m²)", y = "Count",
       title = "Distribution of Energy Efficiency target (heating load)")
```

All 768 target values satisfy `y > 0` (confirmed by `stopifnot` above).

::: {.callout-note}
Features X6 (Orientation, values 2–5) and X8 (Glazing Area
Distribution, values 0–5) are ordinal. The median absolute deviation
of Y1 confirms moderate spread suitable for MAPE evaluation.
:::

## 4 Experimental protocol

The benchmark follows a 30-seed protocol: for each seed in 1–30, the
768 observations are randomly split into 80% training (≈ 614 rows) and
20% test (≈ 154 rows). Hyperparameters are selected via 5-fold
cross-validation on the training set, minimizing each model's native CV
metric. Grid sizes are 48 combinations for SVR-MAPE and Log-SVR; 96
for SVR-MAPE+SK (the symmetry type `a ∈ {-1, 1}` is jointly tuned with
the cost, margin, and bandwidth); 16 for the LS-RMSPE variants and
ε-SVR (MSE); 32 for LS-RMSPE+SK (`a` jointly tuned with cost and
bandwidth); up to 6 for Random Forest; 8 for XGBoost; and no tuning for
Linear Regression, WLS, and Quantile Regression. Six metrics are recorded on the held-out
test set: MAPE, RMSPE, MAAPE, MASE, MSE, and R². Prior to fitting,
predictor features are standardized to zero mean and unit variance
using statistics computed on the training fold only; test-set features
are scaled with the same parameters to prevent data leakage. MASE uses
the lag-1 naive denominator; because the Energy Efficiency data has no
temporal ordering, the denominator depends on row order within each
split and should be interpreted accordingly. Quantile regression (b7a)
could not be fitted on this dataset due to near-singular design matrix
caused by the ordinal features X6 and X8; results for b7a are reported
as NA across all seeds.

## 5 Run experiment

```{r}
#| label: experiment

results_file <- "case-studies/results/energy-efficiency-results.csv"

if (!file.exists(results_file)) {
  results <- run_experiment(X, y,
                            dataset_name = "energy_efficiency",
                            seeds        = 1:30,
                            verbose      = TRUE)
  write_csv(results, results_file)
} else {
  results <- read_csv(results_file, show_col_types = FALSE)
}

cat("Rows loaded:", nrow(results), "\n")
cat("Expected:   ", 30 * 13, "\n")
cat("NA count (b7a expected):", sum(is.na(results$MAPE)), "\n")
```

## 6 Results

### 6.1 Summary table

```{r}
#| label: summary-table

summary_df <- summarise_results(results)
```

```{r}
#| label: summary-table-display
#| echo: false

options(knitr.kable.NA = "—")

fmt_ci <- function(mn, lo, hi) {
  sprintf("%.2f [%.2f, %.2f]", mn, lo, hi)
}

bold_min <- function(x) {
  idx <- which.min(as.numeric(sub(" .*", "", x)))
  x[idx] <- cell_spec(x[idx], bold = TRUE)
  x
}

bold_max <- function(x) {
  idx <- which.max(as.numeric(x))
  x[idx] <- cell_spec(x[idx], bold = TRUE)
  x
}

tbl <- summary_df |>
  transmute(
    Model = label,
    MAPE  = fmt_ci(MAPE_mean,  MAPE_lower,  MAPE_upper),
    RMSPE = fmt_ci(RMSPE_mean, RMSPE_lower, RMSPE_upper),
    MAAPE = fmt_ci(MAAPE_mean, MAAPE_lower, MAAPE_upper),
    MSE   = sprintf("%.2f",  MSE_mean),
    R2    = sprintf("%.4f",  R2_mean)
  ) |>
  mutate(
    MAPE  = bold_min(MAPE),
    RMSPE = bold_min(RMSPE),
    MAAPE = bold_min(MAAPE),
    MSE   = bold_min(MSE),
    R2    = bold_max(R2)
  )

knitr::kable(
  tbl,
  format  = "html",
  escape  = FALSE,
  caption = paste(
    "Values in brackets are 95% percentile bootstrap confidence",
    "intervals over 30 random train/test splits.",
    "b7a (Quantile Regression) could not be fitted due to",
    "near-singular design matrix (ordinal features X6, X8);",
    "results excluded from statistical comparison."
  )
) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                            full_width = FALSE)
```

### 6.2 Box plots

```{r}
#| label: boxplots
#| fig-width: 12
#| fig-height: 8

make_bp <- function(metric, ylab) {
  results |>
    select(abbrev, family, value = all_of(metric)) |>
    filter(!is.na(value)) |>
    mutate(abbrev = fct_reorder(abbrev, value,
                                .fun = median)) |>
    ggplot(aes(x = abbrev, y = value, fill = family)) +
    geom_boxplot(outlier.size = 0.8, linewidth = 0.4) +
    scale_fill_manual(values = c("psvr"     = "#2c7bb6",
                                 "baseline" = "#d7191c"),
                      name = NULL) +
    labs(x = NULL, y = ylab, title = metric) +
    theme_bw(base_size = 11) +
    theme(axis.text.x     = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
}

p_mape  <- make_bp("MAPE",  "MAPE (%)")
p_rmspe <- make_bp("RMSPE", "RMSPE (%)")
p_maape <- make_bp("MAAPE", "MAAPE (%)")
p_r2    <- make_bp("R2",    "R²")

(p_mape + p_rmspe) / (p_maape + p_r2) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### 6.3 Statistical comparison

```{r}
#| label: wilcoxon

results_no_qr <- results |>
  filter(!(model_id == "b7a" & is.na(MAPE)))

wilcox_df <- wilcoxon_vs_best(results_no_qr, metric = "MAPE")
```

```{r}
#| label: wilcoxon-table
#| echo: false

wilcox_df |>
  left_join(
    MODEL_LABELS |> select(id, ref_label = label),
    by = c("reference" = "id")
  ) |>
  select(
    Model                = label,
    `Reference baseline` = ref_label,
    `p-value`            = p_value,
    Significance         = signif
  ) |>
  mutate(`p-value` = sprintf("%.4f", `p-value`)) |>
  knitr::kable(
    caption = paste("Paired Wilcoxon signed-rank test vs.",
                    "best baseline on MAPE. b7a excluded (NA results).")
  )
```

The best baseline in this study is
`r MODEL_LABELS$label[MODEL_LABELS$id == wilcox_df$reference[1]]`.
Of the six `psvr` models,
`r sum(wilcox_df$model_id %in% c("m1","m2","m3a","m3b","m4a","m4b") & wilcox_df$p_value < 0.05, na.rm = TRUE)`
achieve a statistically significant improvement over that baseline at
α = 0.05. Significance codes: \*\*\* p < 0.001, \*\* p < 0.01,
\* p < 0.05, ns = not significant.

## 7 Discussion

Energy Efficiency has moderate target spread (range 6–43 kWh/m²),
making it a favorable dataset for MAPE-optimized models; `psvr` models
achieve competitive MAPE, benefiting from the smooth energy-load
response surface produced by these building simulations.

Ordinal features (X6, X8) do not affect kernel or tree-based methods
but prevent Quantile Regression from fitting — this illustrates a
practical advantage of kernel methods over linear quantile regression
in mixed-type feature spaces. Symmetric kernel models show similar
behavior to the Boston Housing and Diabetes results: the dataset has
no natural geometric symmetry in feature space, so the symmetric
kernel term introduces noise rather than useful inductive bias, and
non-symmetric variants are generally preferable.

As expected, R² is high for all models on this simulation-derived
dataset, where the relationship between architectural features and
heating load is relatively smooth and deterministic. `psvr` models
remain competitive on both MAPE and R² here, unlike the noisier
real-world datasets.

**Practical recommendation:** use SVR-MAPE or LS-RMSPE for building
energy load prediction when percentage-error compliance metrics are
required; kernel methods have an additional practical advantage over
linear quantile regression when ordinal features are present.

## 8 Downloadable files

```{r}
#| label: downloads

xfun::embed_file(
  "case-studies/experiment_helpers.R",
  text = "Download experiment helper script (.R)"
)

xfun::embed_file(
  "case-studies/results/energy-efficiency-results.csv",
  text = "Download full results (30 seeds x 13 models x 6 metrics)"
)
```
