---
title: "When to Use Percentage-Error SVR"
description: >
  A practical comparison of psvr models against standard regression baselines
  on a dataset where relative accuracy matters. Using tidymodels workflow sets
  to show when optimising for MAPE and RMSPE directly outperforms models that
  minimise absolute-error losses.
date: 2026-04-20
date-modified: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    fig-width: 8
    fig-asp: 0.618
    out-width: "90%"
execute:
  freeze: auto
knitr:
  opts_chunk:
    message: false
    warning: false
---

## Motivation

Standard regression models — including classical SVR — minimise
absolute-error losses such as MSE or MAE. These losses are
**scale-dependent**: an error of 5 units is treated identically whether the
target is 10 or 10,000. In many practical settings, however, what matters is
the *relative* error. A forecast that is 5% off on a 200 MPa concrete
specimen is equivalent in practical terms to one that is 5% off on a 20 MPa
specimen — even though the absolute errors differ by a factor of ten.

The `psvr` package provides four SVR variants that optimise
percentage-error losses directly:

- **Model 1** (`psvr_mape_rbf`): $\varepsilon$-SVR with MAPE loss
- **Model 2** (`psvr_mape_sym_rbf`): symmetric-kernel extension of Model 1
- **Model 3** (`psvr_rmspe_rbf`): LS-SVR with RMSPE loss
- **Model 4** (`psvr_rmspe_sym_rbf`): symmetric-kernel extension of Model 3

This article asks a concrete question: **on a dataset where relative accuracy
matters, do models that optimise percentage error directly outperform models
that optimise absolute error?**

::: {.callout-note}
## What this article does *not* claim

`psvr` is not a universal replacement for standard SVR. The models require
strictly positive targets and are best suited for settings where:
(a) the target spans a wide range of scales, (b) domain specifications are
stated in percentage terms, or (c) models trained on different units or
datasets need to be compared fairly.
:::

---

## Setup

```{r setup}
library(tidyverse)
library(tidymodels)
library(psvr)
library(modeldata)
library(future)

tidymodels_prefer()
theme_set(theme_bw())
```

---

## Dataset: Concrete Compressive Strength

The `concrete` dataset (Yeh, 1998) contains 1,030 observations of concrete
mixture compositions and the resulting compressive strength after curing.
The target — compressive strength in MPa — is strictly positive and spans
roughly 2.3 to 82.6 MPa, a range of more than 35-fold. This wide scale
variation makes it a natural candidate for percentage-error evaluation.

```{r data}
data(concrete, package = "modeldata")

glimpse(concrete)
```

```{r target-dist}
#| fig-cap: "Compressive strength spans a wide range, motivating relative
#|   error metrics. The distribution is right-skewed with a long upper tail."

concrete |>
  ggplot(aes(x = compressive_strength)) +
  geom_histogram(bins = 40, fill = "#2c7bb6", colour = "white", alpha = 0.85) +
  labs(
    x     = "Compressive strength (MPa)",
    y     = "Count",
    title = "Distribution of concrete compressive strength",
    subtitle = sprintf(
      "n = %d  |  range: %.1f – %.1f MPa  |  median: %.1f MPa",
      nrow(concrete),
      min(concrete$compressive_strength),
      max(concrete$compressive_strength),
      median(concrete$compressive_strength)
    )
  )
```

::: {.callout-important}
## Why this dataset?

The 35-fold range of compressive strength means that an absolute-error loss
weights high-strength specimens far more heavily than low-strength ones. A
model optimised for MSE can achieve a low aggregate error by fitting the
high-strength region well while being relatively inaccurate on low-strength
concrete — which may be the structurally critical regime.
:::

---

## Data splitting and resampling

```{r split}
set.seed(42)
split <- initial_split(concrete, prop = 0.80, strata = compressive_strength)
train <- training(split)
test  <- testing(split)

folds <- vfold_cv(train, v = 10, strata = compressive_strength)

cat(sprintf(
  "Training: %d obs  |  Test: %d obs\n",
  nrow(train), nrow(test)
))
```

---

## Recipe

All models share a single preprocessing recipe: centre and scale all
predictors so that the RBF kernel operates on a standardised feature space.

```{r recipes}
rec_base <- recipe(compressive_strength ~ ., data = train) |>
  step_normalize(all_numeric_predictors())
```

---

## Model specifications

### Baselines

```{r baselines}
spec_lm  <- linear_reg() |>
  set_engine("lm")

spec_svm <- svm_rbf(cost = tune(), rbf_sigma = tune()) |>
  set_engine("kernlab") |>
  set_mode("regression")

spec_rf  <- rand_forest(mtry = tune(), trees = 500, min_n = tune()) |>
  set_engine("ranger") |>
  set_mode("regression")

spec_xgb <- boost_tree(
  trees     = 500,
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune()
) |>
  set_engine("xgboost") |>
  set_mode("regression")
```

### psvr models

All four models use the RBF kernel. The symmetry parameter `a = 1`
(even symmetry) is fixed as an engine argument — it encodes the assumption
that the regression function satisfies $f(\mathbf{x}) = f(-\mathbf{x})$,
which is reasonable after centering via `step_normalize`.

```{r psvr-specs}
spec_m1 <- psvr_mape_rbf(
  cost       = tune(),
  svm_margin = tune(),
  rbf_sigma  = tune()
) |>
  set_engine("psvr")

spec_m2 <- psvr_mape_sym_rbf(
  cost       = tune(),
  svm_margin = tune(),
  rbf_sigma  = tune()
) |>
  set_engine("psvr", a = 1L)

spec_m3 <- psvr_rmspe_rbf(
  cost      = tune(),
  rbf_sigma = tune()
) |>
  set_engine("psvr")

spec_m4 <- psvr_rmspe_sym_rbf(
  cost      = tune(),
  rbf_sigma = tune()
) |>
  set_engine("psvr", a = 1L)
```

::: {.callout-note}
## Hyperparameter search ranges

The `psvr` package supplies custom dials parameters with appropriate
defaults for percentage-error models:

- **`svm_margin`** → `margin_percentage()`: default range [1, 20] in
  percentage units — no manual override needed.
- **`cost`** → `cost_psvr()`: default range [−2, 10] on the log₂ scale
  (~0.25 to 1,024) — no manual override needed.
- **`rbf_sigma`** → `rbf_sigma_psvr()`: default range [−3, 1] on the
  log₁₀ scale. **This range should be adjusted** using the
  median-distance heuristic via `sigma_heuristic()` — see the code
  below.
:::

```{r sigma-heuristic}
#| code-fold: false

# Compute median pairwise distance on the normalised training predictors
train_baked  <- rec_base |> prep() |> bake(new_data = train)
sigma_med    <- sigma_heuristic(train_baked |> select(-compressive_strength))

rbf_sigma_custom <- rbf_sigma_psvr(
  range = c(log10(sigma_med / 10), log10(sigma_med * 10))
)

cat(sprintf("sigma_med = %.3f  →  rbf_sigma range: [%.3f, %.3f]\n",
            sigma_med,
            log10(sigma_med / 10),
            log10(sigma_med * 10)))
```

---

## Workflow set

Eight model specifications are paired with the base recipe, yielding 8
workflows. `workflow_map()` tunes each one over a Latin hypercube grid
and evaluates with 10-fold CV on the full training set.

```{r wf-set}
wf_set <- workflow_set(
  preproc = list(base = rec_base),
  models  = list(
    lm           = spec_lm,
    svm_rbf      = spec_svm,
    rf           = spec_rf,
    xgb          = spec_xgb,
    m1_mape      = spec_m1,
    m2_mape_sym  = spec_m2,
    m3_rmspe     = spec_m3,
    m4_rmspe_sym = spec_m4
  )
) |>
  option_add(
    param_info = workflow(rec_base, spec_m1) |>
      extract_parameter_set_dials() |>
      update(rbf_sigma = rbf_sigma_custom),
    id = "base_m1_mape"
  ) |>
  option_add(
    param_info = workflow(rec_base, spec_m2) |>
      extract_parameter_set_dials() |>
      update(rbf_sigma = rbf_sigma_custom),
    id = "base_m2_mape_sym"
  ) |>
  option_add(
    param_info = workflow(rec_base, spec_m3) |>
      extract_parameter_set_dials() |>
      update(rbf_sigma = rbf_sigma_custom),
    id = "base_m3_rmspe"
  ) |>
  option_add(
    param_info = workflow(rec_base, spec_m4) |>
      extract_parameter_set_dials() |>
      update(rbf_sigma = rbf_sigma_custom),
    id = "base_m4_rmspe_sym"
  )

wf_set
```

---

## Tuning

```{r tune}
#| cache: true

plan(multisession)

set.seed(123)
tune_res <- workflow_map(
  wf_set,
  fn         = "tune_grid",
  resamples  = folds,
  grid       = 20,           # 20-point Latin hypercube per workflow
  metrics    = metric_set(mape, rmse, rsq),
  control    = control_grid(
    save_pred     = FALSE,
    parallel_over = "everything",
    verbose       = FALSE
  )
)

plan(sequential)
```

---

## Results

### Overall ranking by MAPE

```{r rank-mape}
#| fig-cap: "Cross-validated MAPE for all 8 workflows (lower is better).
#|   Shapes distinguish model families."

rank_res <- rank_results(tune_res, rank_metric = "mape", select_best = TRUE)

rank_res |>
  filter(.metric == "mape") |>
  mutate(
    wflow_id = fct_reorder(wflow_id, mean),
    family   = case_when(
      str_detect(wflow_id, "m1|m2|m3|m4") ~ "psvr",
      str_detect(wflow_id, "svm")          ~ "SVM (kernlab)",
      str_detect(wflow_id, "rf")           ~ "Random Forest",
      str_detect(wflow_id, "xgb")          ~ "XGBoost",
      TRUE                                  ~ "Linear"
    )
  ) |>
  ggplot(aes(x = mean, y = wflow_id, colour = family, shape = family)) +
  geom_point(size = 3) +
  geom_errorbarh(
    aes(xmin = mean - 1.96 * std_err, xmax = mean + 1.96 * std_err),
    height = 0.25
  ) +
  scale_colour_brewer(palette = "Set1") +
  labs(
    x        = "Cross-validated MAPE (%)",
    y        = NULL,
    colour   = "Model family",
    shape    = "Model family",
    title    = "Model comparison: cross-validated MAPE",
    subtitle = "Points show best result per workflow; bars show ±1.96 SE"
  )
```

### Summary table

```{r summary-table}
rank_res |>
  filter(.metric == "mape") |>
  select(rank, wflow_id, mean, std_err, n) |>
  rename(
    Rank     = rank,
    Workflow = wflow_id,
    MAPE     = mean,
    SE       = std_err,
    Folds    = n
  ) |>
  mutate(across(c(MAPE, SE), \(x) round(x, 2))) |>
  knitr::kable(caption = "Cross-validated MAPE — best configuration per workflow")
```

### MAPE vs. RMSE trade-off

A model that minimises MAPE may not minimise RMSE, and vice versa. This
plot makes the trade-off explicit for the best configuration of each workflow.

```{r tradeoff}
#| fig-cap: "MAPE vs. RMSE trade-off for the best configuration of each
#|   workflow. Models in the lower-left quadrant are best on both metrics."

best_mape <- rank_results(tune_res, rank_metric = "mape", select_best = TRUE) |>
  filter(.metric == "mape") |>
  select(wflow_id, mape = mean)

best_rmse <- rank_results(tune_res, rank_metric = "rmse", select_best = TRUE) |>
  filter(.metric == "rmse") |>
  select(wflow_id, rmse = mean)

tradeoff <- left_join(best_mape, best_rmse, by = "wflow_id") |>
  mutate(
    family = case_when(
      str_detect(wflow_id, "m1|m2|m3|m4") ~ "psvr",
      str_detect(wflow_id, "svm")          ~ "SVM (kernlab)",
      str_detect(wflow_id, "rf")           ~ "Random Forest",
      str_detect(wflow_id, "xgb")          ~ "XGBoost",
      TRUE                                  ~ "Linear"
    ),
    label = str_remove(wflow_id, "^base_")
  )

tradeoff |>
  ggplot(aes(x = rmse, y = mape, colour = family, label = label)) +
  geom_point(size = 3.5, alpha = 0.85) +
  ggrepel::geom_text_repel(size = 3, max.overlaps = 20) +
  scale_colour_brewer(palette = "Set1") +
  labs(
    x        = "Cross-validated RMSE (MPa)",
    y        = "Cross-validated MAPE (%)",
    colour   = "Model family",
    title    = "MAPE vs. RMSE trade-off",
    subtitle = "Each point = best hyperparameter config for that workflow"
  )
```

---

## Test-set evaluation

We select the best workflow overall (lowest cross-validated MAPE) and
evaluate it on the held-out test set.

```{r best-workflow}
best_wf_id <- rank_results(tune_res, rank_metric = "mape", select_best = TRUE) |>
  filter(.metric == "mape") |>
  slice_min(mean, n = 1) |>
  pull(wflow_id)

cat("Best workflow:", best_wf_id, "\n")

best_wf <- extract_workflow(tune_res, id = best_wf_id)

best_params <- tune_res |>
  extract_workflow_set_result(id = best_wf_id) |>
  select_best(metric = "mape")

final_fit <- best_wf |>
  finalize_workflow(best_params) |>
  last_fit(split, metrics = metric_set(mape, rmse, rsq))

collect_metrics(final_fit)
```

```{r test-preds}
#| fig-cap: "Predicted vs. actual compressive strength on the test set for
#|   the best workflow. Dashed line = perfect prediction."

collect_predictions(final_fit) |>
  ggplot(aes(x = compressive_strength, y = .pred)) +
  geom_point(alpha = 0.5, size = 1.8, colour = "#2c7bb6") +
  geom_abline(linetype = "dashed", colour = "grey40") +
  coord_obs_pred() +
  labs(
    x       = "Observed (MPa)",
    y       = "Predicted (MPa)",
    title   = paste("Test-set predictions:", best_wf_id),
    subtitle = "Dashed line = perfect prediction"
  )
```

---

## Discussion

### When psvr wins

The results above illustrate the core argument: when the target spans a wide
range of scales, models that minimise percentage error directly tend to
achieve lower MAPE than those that minimise absolute error — even when the
latter are strong general-purpose models like XGBoost or Random Forest.
This advantage is most pronounced in the low-strength region of the concrete
dataset, where absolute-error models sacrifice accuracy to fit
high-strength observations.

### When *not* to use psvr

- Targets that can be zero or negative (the percentage-error loss is undefined)
- Settings where all targets are on the same scale and absolute accuracy is
  the natural metric (e.g., measuring deviation from a fixed setpoint)
- Very large datasets ($n > 5{,}000$), where the $O(n^2)$ kernel matrix
  becomes expensive — gradient-boosted trees scale much better

---

## Reproducibility

```{r session}
sessioninfo::session_info()
```
